---
title: "MACS 30500 LECTURE 11"
author: "Topics: Functions."
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: r
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
# generate CSS file
library(xaringanthemer)
style_duo_accent(
  primary_color = "#011f4b",
  secondary_color = "#bbd6c7",
  inverse_header_color = "#222222",
  black_color = "#222222",
  header_font_google = xaringanthemer::google_font("Atkinson Hyperlegible"),
  text_font_google = xaringanthemer::google_font("Atkinson Hyperlegible"),
  code_font_google = xaringanthemer::google_font("Source Code Pro"),
  base_font_size = "24px",
  code_font_size = "20px",
  # title_slide_background_image = "https://github.com/uc-dataviz/course-notes/raw/main/images/hexsticker.svg",
  # title_slide_background_size = "contain",
  # title_slide_background_position = "top",
  header_h1_font_size = "2rem",
  header_h2_font_size = "1.75rem",
  header_h3_font_size = "1.5rem",
  extra_css = list(
    "h1" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    "h2" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    "h3" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    ".tiny" = list("font-size" = "70%"),
    ".small" = list("font-size" = "90%"),
    ".midi" = list("font-size" = "150%"),
    ".tiny .remark-code" = list("font-size" = "70%"),
    ".small .remark-code" = list("font-size" = "90%"),
    ".midi .remark-code" = list("font-size" = "150%"),
    ".large" = list("font-size" = "200%"),
    ".xlarge" = list("font-size" = "600%"),
    ".huge" = list(
      "font-size" = "400%",
      "font-family" = "'Montserrat', sans-serif",
      "font-weight" = "bold"
    ),
    ".hand" = list(
      "font-family" = "'Gochi Hand', cursive",
      "font-size" = "125%"
    ),
    ".task" = list(
      "padding-right" = "10px",
      "padding-left" = "10px",
      "padding-top" = "3px",
      "padding-bottom" = "3px",
      "margin-bottom" = "6px",
      "margin-top" = "6px",
      "border-left" = "solid 5px #F1DE67",
      "background-color" = "#F3D03E"
    ),
    ".pull-left" = list(
      "width" = "49%",
      "float" = "left"
    ),
    ".pull-right" = list(
      "width" = "49%",
      "float" = "right"
    ),
    ".pull-left-wide" = list(
      "width" = "70%",
      "float" = "left"
    ),
    ".pull-right-narrow" = list(
      "width" = "27%",
      "float" = "right"
    ),
    ".pull-left-narrow" = list(
      "width" = "27%",
      "float" = "left"
    ),
    ".pull-right-wide" = list(
      "width" = "70%",
      "float" = "right"
    ),
    ".blue" = list(color = "#2A9BB7"),
    ".purple" = list(color = "#a493ba"),
    ".yellow" = list(color = "#f1de67"),
    ".gray" = list(color = "#222222")
  )
)

source(here::here("R", "slide-opts.R"))
```

```{r pkgs, include = FALSE, cache = FALSE}
library(tidyverse)
library(palmerpenguins)
#library(here)
#library(countdown)

set.seed(1234)
theme_set(theme_minimal(base_size = rcis::base_size))
```

class: inverse, middle

## Agenda

**Key concepts**
  * Using existing functions
  * Writing your own function
  * Practice 1
  
**More concepts**
  * The `return` statement 
  * `stop()`
  * Anonymous functions
  * Practice 2
  
Next lecture, we conclude this topic by introducing a few more concepts! 
  
<!-- MOVE TO NEXT TIME
  * Writing more complex functions: example
  * Functions with more than one argument
  * Use functions to organize your code
  * Documenting your function
  * Functions and `{{ }}`
  * Practice 3

AFTER CLASS NOTES SUMMER 2024
Online this summer it took 90 min to do part 1 and part 2a (including the second set of practice exercises); moved everything after that to the next lecture 

practice 1: budget 8-10 min to do it and 3-4 minutes to review (about 14 min)

practice 2: budget 10-12 min to do it and 3-4 minutes to review (about 18 min) 

total at least 30 min of practice and about 45 min of explanation 

--> 

---

class: inverse, middle

## Functions key concepts

From using to writing functions:

  * Using existing functions
  * Writing your own function
  * Practice Part 1

---

### Using existing functions

Functions are all over R! We have been **using functions** from day 1 of this class. For example:
```{r}
a <- c(1:5)
a
sum(a)
mean(a)
```


---

### Using existing functions
 
A function: 

* is a block of code that accomplishes a task

* has a name that typically describes the task, like `sum()` or `mean()`

* takes one or more inputs (e.g., arguments), processes it, and returns one or more outputs

---

### Using existing functions

To **use** a function, we need to **know its arguments and pass them correctly.** 

In the console, type `help()` with the function's name in parenthesis to learn more about a function behavior and its arguments:
```{r}
help(sum)
help(mean)
```

<!-- mean(c(100, 1, 2, 3, 4, 5, 100, trim = 0.1)) -->

--

Some functions come from base R, like those above.

Other functions belong to packages, like `summarize()` or `mutate()` which come from `dplyr`

<!-- illustrate help(mutate) but load tidyverse fist --> 

---

### Writing custom functions

Besides **using** functions created by others, we can also **write our own functions from scratch!**

When we write our own function we need to provide the following:

* **name**: something that describes the task that the function performs
* **arguments (input)**: at least one data structure (e.g., value, vector, etc.)
* **body**: self-contained lines of code that manipulate the input, defining what the function does
* **return (output):** statement that returns the result of the function's calculations

Syntax:
```
name <- function(arg1, arg2, ...) { 
    result <- body with code that does things using arg1, arg2, etc.
    return(result)
    }
```

<!--
Notice: 
* we can assign the function to a named object like any other object
* typically the body is multiple lines with multiple intermediate variables
-->

---

### Writing custom functions

For example, rather than using the built-in `sum()` function, we could **write our own sum function** to calculate the sum of the elements of a numeric vector.

Syntax:
```
name <- function(arg1, arg2, ...) { 
    result <- body with code that does things using arg1, arg2, etc.
    return(result)
    }
```

Our sum function: 
```{r}
my_sum <- function(vector) {
  total <- 0
  for (element in vector) {
    total <- total + element
  }
  return(total)
}
```

<!-- LET'S STOP A SECOND AN UNPACK THIS CODE! DO THIS ON WORKBENCH -->

```{r, eval = FALSE, echo = FALSE}
# do not evaluate, do not display

# define a vector
x <- c(2,1,5,2)

# our goal
sum(x)

# the math
sum_x = 2 + 1 + 5 + 2
#mean_x = (1 + 2 + 3 + 4) / 4 

# how we do this programmatically with code for this specific vector?
total <- 0
for (element in x) {
  #print(element)
  total <- total + element
  #print(total)
}
total

# now that we know it works let's put it into a function 
# and test it out with other vectors
my_sum <- function(vector) {
  total <- 0
  for (element in vector) {
    total <- total + element
  }
  return(total)
}

# test with vector x
my_sum(x)

# test with another vector 
y <- c(4,3,7)
my_sum(y)

# or looping over vector indexes (vs. element) generally better
my_sum <- function(vector) {
  total <- 0
  for (i in seq_along(vector)) {
    total <- total + vector[i]
  }
  return(total)
}

```

---

### Writing custom functions

To use a function, we need to **call it** with specific values. Our custom function `my_sum()` takes one argument, `vector`. When we call the function, this will be a concrete vector, such as `x`, `y`, etc. 

```{r}
x <- c(2,1,5,2)
my_sum(vector = x)
```

```{r}
y <- c(4,3,7)
my_sum(y)
```


<!-- SKIP 
````
#j <- c("1", 6, 6)
#my_sum(j)

my_sum <- function(vector) {
  
  # convert vector to numeric, if possible
  # R coerces numeric strings like "1" to numeric values 1 (try if this was "one" )
  vector <- as.numeric(vector)
  #print(vector)
  #print(class(vector))
  
  # same code as above
  total <- 0
  for (v in vector) {
    total <- total + v
  }
  return(total)
}
```
-->

---

### Writing custom functions

Notice a even better way to write our `my_sum` custom function is by looping over the vector indexes vs its elements: 

```
# looping directly over elements
my_sum <- function(vector) {
  total <- 0
  for (element in vector) {
    total <- total + element
  }
  return(total)
}


# looping over indexes (better)
my_sum <- function(vector) {
  total <- 0
  for (i in seq_along(vector)) {
    total <- total + vector[i]
  }
  return(total)
}
```

---

### Advantages of writing custom functions

Use custom-written functions to...

* reduce repetitive code and chances for mistakes

* reuse code

* organize code (e.g., one function imports the data, another cleans the data, etc.)


<!--
Can do with a loop
```
v <- list(a, b)
output <- vector(mode = "list", length = length(v))

for (i in seq_along(v)) {
  output[[i]] <- (sum(v[[i]]/length(v[[i]])))
  #output[[i]] <- (mean(v[[i]]))
}
output
```
-->

---

### Things to remember when writing functions: name

Use **unique names** 
* if your function name matches the name of an existing R function, your function will replace the existing function for your current session (e.g., write `my_sum`, do not re-write `sum`)
* unless there is a specific reason (e.g. learning purposes, etc.), do not create a new function like `my_sum()` when R already has a built-in `sum()` function you can use

Use **informative names**: 
* the name should suggest what the function does 
* it should be short
* avoid reserved words like `if`, `else`, `for`, `function`, etc. To see the full list of reserved words type `help(reserved)` in the Console

---

### Things to remember when writing functions: variables scope

Variables defined inside a function are **not available outside it.** Their scope lies within and is limited to the function itself:

```{r, error = TRUE}
my_sum <- function(input_vector) {
  total <- 0
  for (element in input_vector) {
    total <- total + element
  }
  return(total)
}

my_sum(c(1,2,3))
total
```

---

class: inverse, middle

## Practice writing functions. Part 1

### Download today's class materials from the website. Complete questions 1 and 2.

---

class: inverse, middle

## More concepts

  * The `return` statement 
  * `stop()`
  * Anonymous functions
  * Practice 2
  
---

### The role of the return statement(s)

If you do not write a `return` statement, your output will be the last statement in your code:

```{r}
my_sum <- function(input_vector) {
  total <- 0
  for (element in input_vector) {
    total <- total + element
  }
  total
}

my_sum(c(1,2,3))
```

--

This works, but you might want to explicitly use `return` statements to enhance code clarity and readability. Remember that any code after `return` will be ignored. Let's see an example...

---

### Write multiple return statements

Use multiple returns using `if-else`:

```{r}
check_number <- function(x) {
  if (x > 0) {
    return("positive")
    }
  else if (x < 0) {
    return("negative")
    }
  else {
    return("zero")
    }
}
check_number(1)
```

If x > 0, the function returns "positive" without evaluating rest of the body.

---

### Write one return statement but with multiple outputs

Sometimes you want to return multiple objects and collect them into a list or a vector:

```{r}
my_sum <- function(input_vector) {
  total <- 0
  for (element in input_vector) {
    total <- total + element
  }
  return(list(total, element))
}

my_sum(c(1,2,3))
```

Why is this a list? Can it be a vector in this case? 

<!--

Explain why this is a list, will work as as vector here, but not always depends on type

Try accessing list elements 
s <- my_sum(c(1,2,3))
s[1]
s[2]
s[[1]]
s[[2]]
s[1][[1]]
s[2][[1]]


```
my_mean <- function(vector) {
    total_values <- length(vector)
    result <- (sum(vector/total_values))
    return(list(total_values, result))
}
my_mean(1:10)
```
-->

---

### Using `stop()` when writing a function

Define a function `celsius_to_fahr()` that converts temperatures from Celsius to Fahrenheit using the formula `fahrenheit = (celsius * 9/5) + 32`

```{r}
celsius_to_fahr <- function(celsius) {
  fahr <- (celsius * 9/5) + 32
  return(fahr)
}
```

```{r}
celsius_to_fahr(0)
celsius_to_fahr(-20)
```

---

### Using `stop()` when writing a function

To ensure this function works properly, the argument must be numeric. Otherwise, the conversion between the two temperature scales will not work.

Solution: use `if-else` to verify if the provided argument is numeric; if not, use `stop()` to raise an error and stop the function execution

```{r}
celsius_to_fahr <- function(celsius) {
  if (!is.numeric(celsius)) {
    stop("input must be a number")
  } else {
    fahr <- (celsius * 9/5) + 32
    return(fahr) 
  }
}
```

```{r, error = TRUE}
celsius_to_fahr("zero")
```

<!-- Why use `stop()` and not `print()`?

`stop():` interrupts the execution of the function and throw an error message when a condition is not met 

`print()`: doesnt stop the function execution just print the message
so the first enforces a condition that must be met for the function to execute, and prints a message (which is optional); the other is only a message
-->

---

### Anonymous functions 

<!-- You won't define anonymous or unnamed functions a lot, but you need to know they exist and recognize their syntax.-->

Anonymous functions are **functions without a name**.

**When to use anonymous functions?**

* when the function is short and you need it only once 

* when you use it in conjunction with other functions, such as those from the `apply()` family

In all other cases: do not use them, write the function explicitly instead!

---

### Anonymous functions

Imagine we have the following function:
```{r}
add <- function(x) {
  x + 3
}
add(2)
```

--

Re-write it as an anonymous function. Note: one line, no name, and call it with `()`:
```{r}
(function(x) {x + 3}) (2)
```

--

Often anonymous functions are written without the `{}`, like this:
```{r}
(function(x) x + 3)(2)
```

<!--
Here's an anonymous function for calculating the mean of a vector `x`. In the following example, the input `x` to the function is each element of the list `l`.
```{r}
l <- list(1:5, 5:7)
lapply(l, FUN = function(x){sum(x)/length(x)})
```

see http://adv-r.had.co.nz/Functional-programming.html#anonymous-functions
-->

---

### Anonymous functions: with `sapply()` and `lapply()`

Same function as before but we pass a vector rather than a single number:
```{r}
add <- function(x) {
  x + 1
}
add(c(1,2,3))

```

To rewrite it as anonymous function, this code won't work:
```{r, error = TRUE}
(function(x) x + 1) (1,2,3)
```

It only works on single values, not separated by commas:
```{r}
(function(x) x + 1) (1:3)
```

---

### Anonymous functions: with `sapply()` and `lapply()`

Instead, we can rewrite it using `sapply()`:
```{r}
sapply(1:3, function(x) x + 1)
sapply(c(1,2,3), function(x) x + 1)
```

* `sapply()` means "simplify apply", turn output to a vector
* `lapply()` means "list apply", turn output to a list

Use them to apply a function to each element of a list or vector. The first turns the output into a vector, the latter leaves it as list.

---

### Anonymous functions: the `purrr` way

The `purrr` package (remember the `map()`, `map_dbl()`, etc. functions from last week? they come from this package!) provides another way to write anonymous functions with the **`~.`** syntax. 

Same function as before that adds 1 to the input
```{r}
add <- function(x) {
  x + 1
}
add(c(1,2,3))
```

Rewrite it as anonymous function in `purrr` (https://purrr.tidyverse.org/reference/map.html)
```{r}
map_dbl(c(1,2,3), ~.x + 1)
```

---

class: inverse, middle

## Practice writing functions. Part 2

### Download today's class materials from the website. Complete questions 3, 4, and 5.

