---
title: "HW03: Wrangling and visualizing data"
date: 2023-10-18T13:30:00-06:00  # Schedule page publish date
type: book
toc: true
draft: false
aliases: ["/hw03-wrangle-data.html"]
categories: []
weight: 30
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Overview

**Due Wednesday, October 25th (11:59 PM).**

The goal of this assignment is to practice wrangling and exploring social science data in a research context. This assignment provides a chance to further practice your `dplyr` and `ggplot2` skills, as well as to work with tidy data and multiple datasets.


# Accessing the `hw03` repository

* Go [at this link](https://classroom.github.com/a/kKwGdgcu) to accept and create your private `hw03` repository on GitHub. Once you do so, your repository will be built in a few seconds. It follows the naming convention `hw03-<USERNAME>`  
* Once the your repository has been created, click on the link you see, which will take you to your repository. 
* Finally, clone the repository to your R Workbench (or your computer if you have installed R there) following the process below.


# Cloning your `hw03` repository

After you have accessed the `hw03` repository (see above), follow the [same steps you completed for `hw01`](/homework/edit-readme/) to clone the repository.


# General workflow

See Homework 1

# Assignment description

This assignment has two separated parts. The first asks you to tidy messy data, the second asks you to work with the supreme court database (recode variables and wrangle data to answer a set of questions).

## PART 1: Tidying messy data

In the `rcis` package, there is a data frame called `dadmom`.

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
library(rcis)

data("dadmom")
dadmom
```

Tidy this data frame so that it adheres to the tidy data principles. Your final product (tidy data frame) should look like this:

```{r, echo = FALSE}
ex <- dadmom %>%
  # use pivot_longer() to collect the relevant columns
  pivot_longer(
    cols = named:incm,
    names_to = c("variable"),
    values_to = "value",
    # name and inc store data in separate column types
    # simplify everything to character columns to avoid an error
    values_transform = as.character
  ) %>%
  # use separate() to split variable into name/inc column and d/m column
  separate(col = variable, into = c("variable", "parent"), sep = -1) %>%
  # now pivot the name/inc variables into separate columns
  pivot_wider(
    names_from = variable,
    values_from = value
  ) %>%
  # ensure income is stored as a numeric column
  # alas there is no way to do this within pivot_wider()
  mutate(inc = parse_number(inc)) 

ex2 <- as.data.frame(ex)
ex2
```

You can accomplish this task using a single or multiple piped operations using only `tidyr` functions. Code that does not use any `tidyr` functions is not acceptable. Check the lectures, readings, and in-class exercises before starting this part.

Once you have tidied the data frame, generate a plot using the exact code provided in your repo. Do not change the code: if you tidied the data frame correctly, you will not have to make any changes to that code. If errors arise, it means the data has not been tidied correctly. 


## PART 2: Wrangling and visualizing messy(ish) data

### Description of the data: Supreme Court Database

The [Supreme Court Database](http://scdb.wustl.edu/) contains detailed information of every published decision of the U.S. Supreme Court since its creation in 1791. It is perhaps the most utilized database in the study of judicial politics.

In the `hw03` repository, you will find two data files:

1. `scdb-case.csv`
1. `scdb-vote.csv`

These contain the exact same data you would obtain if you downloaded the files from the original website, but reformatted to be stored as relational data files. That is, `scdb-case.csv` contains all *case-level* variables, whereas `scdb-vote.csv` contains all *vote-level* variables.

The data is structured in a tidy fashion:

* `scdb-case.csv` contains one row for every case and one column for every variable
* `scdb-vote.csv` contains one row for every vote by a justice in every case and one column for every variable

The current data contain information on every case decided from the 1791-2020 terms.[^terms] There are several ID variables which are useful for other types of research: for our purposes, the only ID variable you need to concern yourself with is `caseIssuesId`. Variables you will want to familiarize yourself with include:

* `chief`
* `dateDecision`
* `decisionDirection`
* `decisionType`
* `declarationUncon`
* `direction`
* `issueArea`
* `justice`
* `justiceName`
* `majority`
* `majVotes`
* `minVotes`
* `term`

{{% callout note %}}
Read the [documentation](http://scdb.wustl.edu/documentation.php) to see how these variables are coded.
{{% /callout %}}


### Task 1: Recode variables

Recode the following variables according to the instructions. You will be using the newly generated variables to answer the questions below. For each variable, show the total observations for each value before and after recoding:

* `decisionDirection`: this variable takes four values (1 conservative, 2 liberal, 3 unspecifiable, and NA), as explained in the [documentation](http://scdb.wustl.edu/documentation.php?var=decisionDirection#norms). Recode it to take three values: 0,1,NA (0 conservative, 1 liberal, and NA as a sum of NA and unspecifiable)
* `majority`: this variable takes two values (1 dissent, 2 majority), as explained in the [documentation](http://scdb.wustl.edu/documentation.php?var=majority). Recode it so dissent becomes 0 and majority becomes 1
* `direction`: this variable takes two values (1 conservative, 2 liberal), as explained in the [documentation](http://scdb.wustl.edu/documentation.php?var=direction). Recode it so conservative becomes 0 and liberal becomes 1
* convert the `chief` and `justiceName` variables to a factor (with chief and justice in chronological order); please notice, these variables are stored into two different data sources


### Task 2: Answer the following questions 

For each question, provide 1-2 paragraphs of written interpretation of your results. Graphs and/or tables alone will not be sufficient to answer these questions. The interpretation should be descriptive (e.g., this graph shows x and y) and substantive (what do the results show, what can we infer or conclude from the graph, etc., see EDA lecture on this). Add your description as text, not as an R comment, and place it after your code and results.

Pay careful attention to the unit of analysis required to answer each question. Some questions only require case-level variables, others only require vote-level variables, and some may require combining the two data frames together. Be sure to choose an appropriate relational join function as necessary.

The first three questions are required, the fourth question is a bonus/optional question.
 
 
1. What percentage of cases in each term are decided by a one-vote margin (i.e. 5-4, 4-3, etc.)?

2. How often the justices who are currently serving on the Supreme Court, have voted in the conservative direction in cases of criminal procedure, civil rights, economic activity, and federal taxation?

3. For each term he served on the Court, in what percentage of cases was Justice Antonin Scalia in the majority?

4. Optional/Bonus question. In each term, what percentage of cases were decided in the conservative direction? And in the liberal direction? 


# Submit the assignment

To submit the assignment, follow these steps:

* First push to your repository the last version of your assignment before the deadline. Make sure to stage-commit-push the following files:

  - `dadmom.Rmd`: main file you will add your code to
  - `dadmom.md`: you will generate this file from the .Rmd by knitting it
  - `scouts.Rmd`: main file you will add your code to
  - `scotus.md`: you will generate this file from the .Rmd by knitting it
  - the folders with all the graphs you generated in your .Rmd files

* When you are ready to submit, copy your repository URL (e.g. https://github.com/css-fall23/hw3-brinasab) and submit it on Canvas under HW03 before the deadline. Do not submit files on Canvas, we only need the link to your repository 

* As part of your submission, in the Rmd for this homework:
  * write 1-2 paragraphs of reflections about the homework
  * list the first and last name of eventual collaborators with whom you worked with to complete this assignment


# Assessment

All homework assignments are evaluated using [this general rubric](/faq/homework-evaluations/). Make sure to check it. 

Below are further guidelines for this specific homework to help you assess your work before submitting it.

*Needs improvement:* Doesn't complete all parts of the assignment. Code contains errors and/or is not clearly written and/or not properly documented. Variables are not properly recoded. Plots are incomplete, or use the same type of plot for each graph, or doesn't use plots appropriate for the variables being analyzed. Interpretation is lacking or incomplete. Minimal record of commits other than the final push to GitHub (we expect students to commit often, and commit messages should be self-explanatory and informative).

*Good:* Solid effort. Hits all the elements. Finished all components of the assignment with only minor deficiencies. Easy to follow (both the code and the output). Shows sufficient understanding of the packages needed for the assignment. Attempt to answer bonus question.

*Excellent:* Finished all components of the assignment correctly with well-written, efficient, and well-documented code (both self-documented and with additional comments as necessary). Variables are correctly re-coded, possibly with the use of different approaches. Graphs and tables are properly chosen and labeled. Visuals are well-crafted. Analysis and interpretation of results are clear, elaborated, and easy to follow. Answers show solid understanding of the packages needed for the assignment. The bonus question is answered correctly and/or the other questions are answered going above/beyond what expected and showing excellent mastery of the materials. Repo displays history of multiple commits to back up and show a progression in the work.

[^terms]: Terms run from October through June, so the 2020 term contains cases decided from October 2020 - June 2021.


# Acknowledgments
```{r child = here::here("R", "_ack_ben.Rmd")}
```