---
date: "2022-09-25T00:00:00-05:00"
draft: false
weight: 40
title: "Assignments & Assessment"
toc: true
type: book
aliases: ["/hw00_homework_grading.html", "/hw00_github_rubric.html"]
---

## Assignments

**Required work**: Students will complete a series of programming assignments linked to class materials. Specifically, there will be six homework assignment and a final project. (HW01) is evaluated on a Pass/Fail basis. Subsequent homework assignments (HW2 to HW6) and the final project are graded using a point system and a rubric, described below. See the [Homework](https://computing-soc-sci.netlify.app/homework/) section of this website for more details on each assignment. Assignments will initially be detailed and come with starter code or an initial version of the program where you need to fill in the blanks to make it work. As the course moves on and your skills become more developed, less help upfront will be provided. 

**Deadlines and late work**: All assignments are due at 11:59 PM Chicago time (submit the link to your GitHub repo on Canvas). There will be a 15-minute grace period. After that, late submissions will be penalized by half-a-point for every day (24h) they are late. For example, if your score for a given assignment is 14.5 points and you turn it in one day late, your final score will be 14 points. Summer courses are intense and proceed at a fast speed, submitting late work will undermine your learning. 

**Missed Assignments**: We expect students to submit all assignments. A missed assignment is worth 0 points (0%) and, as such, will negatively affect your final grade. Please reach out if you miss an assignment due to extraordinary circumstances (illness or emergency). I may grant additional extensions on a case-by-case basis. Whenever possible, inform me before the deadline. Things like having a heavy workload in a given week or starting a given assignment too late do not qualify as extraordinary circumstances. 

**Incomplete Assignments**: If more than 25% of the assignment is missing, we won't grade it and will treat it as a missed assignment (see point above). Otherwise, we will grade it with a penalty for each element of the rubric (e.g., if you score "Good" on "Reproducibility," you will gain "Needs Improvement," etc.). 

**Plagiarism**: While students are encouraged to assist one another in debugging programs and solving problems, and are allowed to use online resources it is imperative that students learn how to write code, debug, etc. for themselves. That is, students need to understand, write, and submit their own work. We will run random checks, and if we detect plagiarized code, the whole assignment will be graded as 0. Read our [Plagiarism policy](https://computing-soc-sci.netlify.app/faq/course-obj-expectations/#plagiarism-and-academic-integrity) for details.

<!--
* You can re-do one assignment of your choice and we will re-grade it. The re-do can be submitted at any time before the last day of class and must include a narrative that explains why you are re-doing the assignment and what you have improved (e.g., "I did not understand functions that well, especially this and that. Now...")
-->

## Evaluation philosophy

I do not think that grades are the best way to support learning and demonstrate progress. On the contrary, I find that numeric scores on assignments tend to cause students anxiety if they do not achieve a perfect 100%. However, this university, and most education systems more generally, require that instructors assign each student a letter grade at the end of the course.

To alleviate some of the anxiety and improve the transparency,  consistency, and fairness of grading, we use a detailed rubric (below) to evaluate all submitted work. Most importantly, we encourage students to reach out and ask for additional feedback on what has been done well and what can be improved. 

Remember that a numeric score does not evaluate your worth or intellectual abilities. It is simply an evaluation of a specific piece of submitted work. 


<!--
As such, I do not assign numeric scores in this class. All homework assignments are evaluated using the grading rubric below. Final grades are calculated as the cumulative performance across all homework assignments. 

I find that numeric scores on assignments tend to cause students grading anxiety if they do not achieve a perfect 100%. Nor does the numeric score convey specific feedback on what the student has done well on an assignment and on areas for improvement.

As such, I do not assign numeric scores in this class. All homework assignments are evaluated using the grading rubric below. Final grades are calculated as the cumulative performance across all homework assignments. 

Failure to complete the two weekly [peer evaluation assignments](/faq/peer-evaluations/) causes a minor deduction in the final grade.
-->

## Rubric 

Your assignments will be evaluated using the rubric below (also posted on Canvas). In addition, each assignment will have specific guidelines that further explain our expectations. These will be posted in the assignment itself. 

Your assignment will score *Excellent*, *Very Good*, *Good*, *Needs improvement*, or *Not sufficient* on the following five categories:

* *Coding style:* coding style pertains to stylistic issues, not to whether your code works. It evaluates how well the code follows the [convention for R code](http://adv-r.had.co.nz/Style.html) and how consistent it is. Examples include clarity of the code to read for a person who has not written it, use of comments (code should never be over-commented, nor under-commented), proper variables names, understandable/logical code organization, etc. See [here](https://www.smashingmagazine.com/2012/10/why-coding-style-matters/) for more and why this matters.

* *Coding strategy:* by coding strategy, we mean the logic of your code (how your code approaches and solves the given problem), its correctness (whether it works), how sophisticated it is, how well it incorporates the tools and techniques we learn in class (e.g., conditionals, functions, general use of tidyverse syntax, etc.), and whether it minimizes repetition and optimize efficiency.

* *Presentation:* in assignments that have data visualization or data presentation components of any sort (tables, graphs, descriptive stats, more sophisticated analyses), presentation evaluates how well these components are executed both graphically and conceptually (your choices of how to present data). In all assignments, presentation will also evaluate your mastery of Markdown syntax. 

* *Achievement:* evaluates your understanding of the concepts/tools required to complete the assignment (how well you master them) and your ability to use and go beyond the basic tools; it praises extraordinary work produced in the assignment.

* *Reproducibility:* evaluates how well the assignment is in compliance with the course convention for submitted work in this course (e.g., whether we can access your repo, you submit all required elements, your use of Markdown and RMarkdown, Git/GitHub, README, etc.), and whether the submitted materials allow to successfully reproduce your work. 

To achieve *Excellent* in a given category, all requirements must be satisfied. For the other categories, it might be that only one (or more than one) requirement is missing. For example, you could achieve *Very Good* on "Coding Style" if you are following all coding conventions and are submitting a well-commented code but the code could benefit from omre refinement or is inconsistent.

Details on how each rubric score ("Excellent", "Very Good", etc.) translates into points and grades for each assignment are provided on Canvas. Generally, "Excellent" translates into A, "Very Good" into A-, "Good" is B, "Needs Improvement" is C, and "Not sufficient" is below C.

Your final grade is the cumulative score of all submitted work. I use the following *grading scheme* to determine your final grade: A (95-100), A- (90-94), B+ (87-89), B (83-86), B- (80-82), C+ (77-79), C (73-76), C- (70-72). Grades below 70 are dealt on a case-by-case basis.

If you plan to take this course for *Pass/Fail*, email me to make a shared plan that meets your needs and is in-line with course expectations. A "Pass" in this course is a C-.


<!--
Topic| Excellent: <br> ✓+ coded as +  | Good: <br> ✓ coded as 0  |Needs improvement: <br> ✓- coded as - |
|-----------| ---------------------- |--------------------------| ---------------------------------------|
|Coding style| Code is clear and well organized. Coding conventions are followed consistently (spaces, variable names, etc.). Code is well-commented and easy to read. |  Code is generally clear but could be more refined and/or is inconsistent. Coding conventions are partly followed. Code is readable but lacks some comments or is overly commented. | Code lacks refinement and/or is inconsistent. Coding conventions minimally or not followed. Code has no comments and little attention is paid to making the code human readable. |
|Coding strategy| Logic behind the code is clear. Complicated problem broken down into sub-problems that are individually much simpler. Code is efficient, correct, and minimizes repetitions. Code uses appropriate concepts learned in class (e.g. use proper data structure, etc.). All code runs correctly and checks for common errors. | Logic is overall clear. Code is mostly correct, but could be edited down to leaner code. Some "hacking" instead of using proper tools (e.g., suitable data structure) and/or efficiency and repetitions could be improved. | Logic is difficult to follow. Code tackles complicated problem in one big chunk. Some parts of the code might be repetitive, for example it could easily be functionalized. Code does not check for errors and/or code does not run.|
|Presentation| Graph(s)/table(s) carefully tuned for desired purpose. One graph/table illustrates one point. Careful styling highlights important features. Full command of Markdown syntax and its components. | Graph(s)/table(s) well chosen, but with a few minor problems (e.g., inappropriate aspect ratios, poor labels, formatting deficiencies, etc.). Appropriate use of Markdown syntax and its components, with some minor deficiencies. | Graph(s)/table(s) poorly chosen to support questions and illustrate findings. Major display problems with graphs, tables and/or Markdown components.|
|Achievement| Assignment has gone beyond what expected and required (e.g., extraordinary outcome, additional tools not explicitly addressed by this course and/or sophisticated use of tools from course).| Tools and techniques from the course are applied competently and somewhat creatively. Chosen task is good, but fairly conservative in ambition.|Assignment does not display the expected level of mastery of the tools and techniques in this course. Chosen task is too limited in scope.|
|Reproducibility| Full compliance with course conventions for submitted work in this course. Access is as easy as possible, complies with reproducibility conventions, code runs, workflow is correct. | Code partially complies with reproducibility conventions and/or code does not run. | Not an earnest effort to reduce friction and comply with conventions and/or code does not run.|

## Template

```
Evaluation
-----------------------------------------------------------------------------
| Topic                       | Excellent | Good         | Needs Improvement |
|-----------------------------|-----------|--------------|------------|
| **Coding style**            |           |              |            |
| **Coding strategy**         |           |              |            |
| **Presentation**            |           |              |            |
| **Achievement**             |           |              |            |
| **Reproducibility**         |           |              |            |

Examples of further remarks:
* Elaborate on above, especially for "Needs improvement"
* Specific praise
* Specific constructive criticism

```

## How do the rubric scores convert to a letter grade?

* You can think at "Good" as being equivalent to a B+. So if you hypothetically earn "Good" on every assignment for every rubric element, at the end of the quarter you would earn a B+. 
* If you earn a combination of "Good" and "Excellent," you are looking at the difference between a B+ and an A-, or an A- and an A (e.g., if the majority of your evaluations are "Excellent" you are closer to an A, if they the majority are "Good" you are closer to a B+).
* If you earn a combination of "Good" and "Needs Improvement" scores, then you might be somewhere between a B+ and a B, or lower.
* The same logic applies if you earn a combination of all three scores: your final grade will depend on how many scores for each category

Historically, the majority of students in the class earn a B+ or higher.

The final grade is calculated as the cumulative performance (sum of all "Excellent", "Good", "Needs Improvement" across all homework assignments), assuming you are submitting all assignments: missed assignments or incomplete work will negatively affect your grade (see above for more info)

## Acknowledgments

```{r child = here::here("R", c("_ack_stat545.Rmd", "_ack_ben.Rmd"))} 
```
-->


<!--
You will complete a series of programming assignments throughout the quarter linked to class materials. Assignments will initially come with starter code, or an initial version of the program where you need to fill in the blanks to make it work. As the quarter moves on and your skills become more developed, less help upfront will be provided. Each assignment will be evaluated by myself or a TA.

Each assignment will be evaluated by myself or a TA, as well as by *two peers*. Peer review is a crucial element to this course, in that by [eating each other's dog food](https://en.wikipedia.org/wiki/Eating_your_own_dog_food) you will learn to read, debug, and reproduce code and analysis. And while I and the TAs are competent users in R, your classmates are not - so make sure your code is [well-documented](#documentation) so that others with basic knowledge of programming and R can follow along and reuse your code. Be sure to read the instructions for [peer review](/faq/peer-evaluations/) so you know how to provide useful feedback.

<!--
As part of this course you will be reviewing, commenting on, and marking other students' assignments. This is a mandatory part of the course: failure to complete peer reviews will result in a mark down on your final grade.

## Expectations for peer reviewer

* Identify three **specific things** your peer did well
* Identify three **specific things** the student could improve upon (and ideally provide a suggested approach or solution)

## How to do peer review well

* Give thoughtful, constructive and considerate comments
* Be specific and concise
* Use [the rubric](/faq/homework-evaluations/) for ideas about criteria to evaluate and comment on
* Try to learn something new and, if you succeed, point that out
* If you can't find anything to praise or that you found helpful, then at least offer some suggestions in a kind way
* See [here](https://help.github.com/articles/reviewing-proposed-changes-in-a-pull-request/) for useful instructions on how to initiate and submit reviews using GitHub's built-in tools
* To ensure reproducibility, you might find it useful to clone your classmate's repo and attempt to run their script(s). If you cannot execute them, then the code is not reproducible. Also be aware your classmates will hold you to a similar standard.

## How to do peer review bad

* Your review is so generic that it's hard to determine which assignment you're reviewing
* Your review is mean
* You can't find anything to praise/learn and yet you don't offer any suggestions either

Performing good peer review is difficult! In graduate school we are taught to criticize and tear down others' work and find the flaws. We need to be better at this and not just criticize, but highlight good aspects and suggest how to improve the work. This is still a habit I am struggling to break, so start working on it now before you leave grad school.

## Acknowledgments

```{r child = here::here("R", "_ack_stat545.Rmd")}
```
-->

