---
date: "2022-09-25T00:00:00-05:00"
draft: false
weight: 40
title: "Homework Rubric"
toc: true
type: book
aliases: ["/hw00_homework_grading.html", "/hw00_github_rubric.html"]
---

## How will I be evaluated?

Students will complete a series of (roughly) weekly programming assignments linked to class materials. 

Assignments will initially come with starter code, or an initial version of the program where you need to fill in the blanks to make it work. As the quarter moves on and your skills become more developed, less help upfront will be provided. 

While students are encouraged to assist one another in debugging programs and solving problems in these assignments, it is imperative students also learn how to do this for themselves. That is, **students need to understand, write, and submit their own work.**


## Evaluation philosophy

The university requires that I assign each student a letter grade at the end of the quarter. However I find that numeric scores on assignments tend to cause students grading anxiety if they do not achieve a perfect 100%. Nor does the numeric score convey specific feedback as to what the student has done well on an assignment, combined with feedback on areas for improvement.

As such, I do not assign numeric scores in this class. All homework assignments are evaluated using the grading rubric below. Final grades are calculated as the cumulative performance across all homework assignments. 

<!--
Failure to complete the two weekly [peer evaluation assignments](/faq/peer-evaluations/) causes a minor deduction in the final grade.
-->

## How do the rubric scores convert to a letter grade?

<!--
One year on my course evaluations a student commented

> [T]he grading policy is locked in some chest somewhere under the ocean

Let's pull back the curtain and demystify how I calculate final grade! 
-->

* In this course, "Satisfactory" is equivalent to a B+. So if you earned "Satisfactory" on every assignment for every rubric element, at the end of the quarter you would earn a B+. 
* If you earn a combination of "Satisfactory" and "Excellent" scores you are looking at the difference between a B+ and an A-, or an A- and an A. 
* If you earn a combination of "Satisfactory" and "Needs Improvement" scores, then you might be somewhere between a B+ and a B or lower.
* Etc. etc.

Historically, the majority of students in the class earn a B+ or higher. 

The final grade is calculated assuming you are submitting all homework assignments: missed assignments or incomplete work will negatively affect your grade. 

We have a total of 8 assignments in the class; each is worth roughly 12.5% of your final grade. A missed assignment is worth 0%.

ADD INFO HERE


## Rubric

*In addition to this general rubric, make sure to consult any specific guidance given in the relevant assignment itself*

Topic| Excellent: <br> ✓+ coded as +  | Satisfactory: <br> ✓ coded as 0  |Needs improvement: <br> ✓- coded as - |
|-----------| ---------------------- |--------------------------| ----------------------|
|Coding style| Student has gone beyond what was expected and required, coding conventions are followed, code is well commented. | Coding style lacks refinement and has some errors, but code is readable and has some comments. | Many errors in coding style, little attention paid to making the code human readable.|
|Coding strategy| Complicated problem broken down into sub-problems that are individually much simpler. Code is efficient, correct, and minimal. Code uses appropriate data structure (list, data frame, vector/matrix/array). Code checks for common errors.  | Code is correct, but could be edited down to leaner code. Some "hacking" instead of using suitable data structure. Some checks for errors. |   Code tackles complicated problem in one big chunk. Code is repetitive and could easily be functionalized. No anticipation of errors. |
|Presentation | Graph(s) carefully tuned for desired purpose. One graph illustrates one point. <br><br> Table(s) carefully constructed to make it easy to perform important comparisons. Careful styling highlights important features. <br><br> Full command of Markdown syntax and its components. | Graph(s) well chosen, but with a few minor problems: inappropriate aspect ratios, poor labels. <br><br> Table(s) generally appropriate but possibly some minor formatting deficiencies. <br><br> Appropriate use of Markdown syntax and its components, but with some minor deficiencies. | Graph(s) poorly chosen to support questions. <br><br> Table(s) with too many, or inconsistent, decimal places. Table(s) not appropriate for questions and findings. <br><br> Major display problems with graphs, tables and/or Markdown components.|
|Achievement, mastery, cleverness, creativity|Student has gone beyond what was expected and required, e.g., extraordinary effort, additional tools not addressed by this course and/or shows sophisticated application of tools from course.|Tools and techniques from the course are applied very competently and, perhaps, somewhat creatively. Chosen task was acceptable, but fairly conservative in ambition.|Student does not display the expected level of mastery of the tools and techniques in this course. Chosen task was too limited in scope.|
|Reproducibility, compliance with course conventions for submitted work|Access is as easy as possible, comply with reproducibility conventions, code runs, workflow is correct. | Code partially complies with reproducibility conventions and/or code does not run. | Not an earnest effort to reduce friction and comply with conventions and/or code does not run.|

## Template

```
Evaluation
----------------------------------------------------
| Topic                       | Excellent | Satisfactory | Needs Improvement |
|-----------------------------|-----------|--------------|------------|
| **Coding style**            |           |              |            |
| **Coding strategy**         |           |              |            |
| **Presentation**            |           |              |            |
| **Achievement, creativity** |           |              |            |
| **Reproducibility**         |           |              |            |

Examples of further remarks:

* Elaborate on above, especially for "Needs improvement"
* Some specific praise
* Specific constructive criticism

```

## Acknowledgments

```{r child = here::here("R", c("_ack_stat545.Rmd", "_ack_ben.Rmd"))} ```




<!--
You will complete a series of programming assignments throughout the quarter linked to class materials. Assignments will initially come with starter code, or an initial version of the program where you need to fill in the blanks to make it work. As the quarter moves on and your skills become more developed, less help upfront will be provided. Each assignment will be evaluated by myself or a TA.

Each assignment will be evaluated by myself or a TA, as well as by *two peers*. Peer review is a crucial element to this course, in that by [eating each other's dog food](https://en.wikipedia.org/wiki/Eating_your_own_dog_food) you will learn to read, debug, and reproduce code and analysis. And while I and the TAs are competent users in R, your classmates are not - so make sure your code is [well-documented](#documentation) so that others with basic knowledge of programming and R can follow along and reuse your code. Be sure to read the instructions for [peer review](/faq/peer-evaluations/) so you know how to provide useful feedback.

<!--
As part of this course you will be reviewing, commenting on, and marking other students' assignments. This is a mandatory part of the course: failure to complete peer reviews will result in a mark down on your final grade.

## Expectations for peer reviewer

* Identify three **specific things** your peer did well
* Identify three **specific things** the student could improve upon (and ideally provide a suggested approach or solution)

## How to do peer review well

* Give thoughtful, constructive and considerate comments
* Be specific and concise
* Use [the rubric](/faq/homework-evaluations/) for ideas about criteria to evaluate and comment on
* Try to learn something new and, if you succeed, point that out
* If you can't find anything to praise or that you found helpful, then at least offer some suggestions in a kind way
* See [here](https://help.github.com/articles/reviewing-proposed-changes-in-a-pull-request/) for useful instructions on how to initiate and submit reviews using GitHub's built-in tools
* To ensure reproducibility, you might find it useful to clone your classmate's repo and attempt to run their script(s). If you cannot execute them, then the code is not reproducible. Also be aware your classmates will hold you to a similar standard.

## How to do peer review bad

* Your review is so generic that it's hard to determine which assignment you're reviewing
* Your review is mean
* You can't find anything to praise/learn and yet you don't offer any suggestions either

Performing good peer review is difficult! In graduate school we are taught to criticize and tear down others' work and find the flaws. We need to be better at this and not just criticize, but highlight good aspects and suggest how to improve the work. This is still a habit I am struggling to break, so start working on it now before you leave grad school.

## Acknowledgments

```{r child = here::here("R", "_ack_stat545.Rmd")}
```
-->

