---
title: "MACS 30500 LECTURE 9"
author: "Topics: Intro to Control Structures. Conditional Statements."
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: r
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
# generate CSS file
library(xaringanthemer)
style_duo_accent(
  primary_color = "#011f4b",
  secondary_color = "#bbd6c7",
  inverse_header_color = "#222222",
  black_color = "#222222",
  header_font_google = xaringanthemer::google_font("Atkinson Hyperlegible"),
  text_font_google = xaringanthemer::google_font("Atkinson Hyperlegible"),
  code_font_google = xaringanthemer::google_font("Source Code Pro"),
  base_font_size = "24px",
  code_font_size = "20px",
  # title_slide_background_image = "https://github.com/uc-dataviz/course-notes/raw/main/images/hexsticker.svg",
  # title_slide_background_size = "contain",
  # title_slide_background_position = "top",
  header_h1_font_size = "2rem",
  header_h2_font_size = "1.75rem",
  header_h3_font_size = "1.5rem",
  extra_css = list(
    "h1" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    "h2" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    "h3" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    ".tiny" = list("font-size" = "70%"),
    ".small" = list("font-size" = "90%"),
    ".midi" = list("font-size" = "150%"),
    ".tiny .remark-code" = list("font-size" = "70%"),
    ".small .remark-code" = list("font-size" = "90%"),
    ".midi .remark-code" = list("font-size" = "150%"),
    ".large" = list("font-size" = "200%"),
    ".xlarge" = list("font-size" = "600%"),
    ".huge" = list(
      "font-size" = "400%",
      "font-family" = "'Montserrat', sans-serif",
      "font-weight" = "bold"
    ),
    ".hand" = list(
      "font-family" = "'Gochi Hand', cursive",
      "font-size" = "125%"
    ),
    ".task" = list(
      "padding-right" = "10px",
      "padding-left" = "10px",
      "padding-top" = "3px",
      "padding-bottom" = "3px",
      "margin-bottom" = "6px",
      "margin-top" = "6px",
      "border-left" = "solid 5px #F1DE67",
      "background-color" = "#F3D03E"
    ),
    ".pull-left" = list(
      "width" = "49%",
      "float" = "left"
    ),
    ".pull-right" = list(
      "width" = "49%",
      "float" = "right"
    ),
    ".pull-left-wide" = list(
      "width" = "70%",
      "float" = "left"
    ),
    ".pull-right-narrow" = list(
      "width" = "27%",
      "float" = "right"
    ),
    ".pull-left-narrow" = list(
      "width" = "27%",
      "float" = "left"
    ),
    ".pull-right-wide" = list(
      "width" = "70%",
      "float" = "right"
    ),
    ".blue" = list(color = "#2A9BB7"),
    ".purple" = list(color = "#a493ba"),
    ".yellow" = list(color = "#f1de67"),
    ".gray" = list(color = "#222222")
  )
)

source(here::here("R", "slide-opts.R"))
xaringanExtra::use_panelset()
```

```{r pkgs, include = FALSE, cache = FALSE}
library(tidyverse)
library(rcis)
library(countdown)
library(modeldata)
library(here)

set.seed(1234)
theme_set(theme_minimal(base_size = 16))
```

class: inverse, middle

# Agenda

  * Control structures: definition 
  
  * Conditional statements
  

<!-- dropped vectors and lists from original version of this lecture
and moved them to the data structure lecture; instead I added here conditional statements and more examples and content on iteration 

IDEA for next time: split this into two

lecture 1: review data structure / base R + conditional statements (and add more in-class practice and how to do conditional statements with a dataframe, use penguins_clean!)

lecture 2: loops slides and demo + loop exercises + maps/across + while loops. When you teach for loops add the break and continue statements (currently not in slides!)

NB: in the demo for loop add data structure, e.g. show how to access columns of a df and their elements with the double and single square brakets 

-->

---

class: inverse, middle

## What are control structures? 

---

### Control structures: definition

All the code we have written so far can be seen as a finite and fixed sequence of commands.

**Control structures** enable us to vary our commands by **controlling their flow of execution**.

They do so by integrating logical conditions into our code, enabling different lines of code to execute under varying conditions.

This approach differs from executing the same code every time, as we have done so far.

---

### Main control structures

* **Conditional statements**: test one or more condition(s) and act on it/them

* **`for` loop**: execute a block of code for a fixed number of times

* **`while` loop**: execute a block of code while a given condition is true (and stops as long as the condition is evaluated as false)

<!-- We will now explain each of these control structures, and do practice exercises, but make sure to review the assigned readings for further details: one is from Chapter 13 “Control Structures” in *R programming for Data Science* and Chapter 21 “Iteration” in *R for Data Science* for details. -->

---

class: inverse, middle

## Conditional statements
 
 * single test with `if-else`
 * multiple tests with `if, else if, else`
 * nested conditionals
 * `ifelse()` or its tidyverse version `if_else()`
 * practice exercises
   

---

### if-else 

if-else tests a condition and acts on it depending on whether the condition is TRUE or FALSE.

Syntax:
```
if (condition to be evaluated) {
    action performed when condition is TRUE
} else {
    action performed when condition is FALSE
}
```
--

Example:

```{r echo = TRUE, results = "hide"}
age <- 14
if (age > 16) {
  print("You can get a driving license")
} else {
  print("You cannot drive")
}
```

<!-- available options for size in descending order are:
Huge > huge > LARGE > Large > large > normalsize > small > footnotesize > scriptsize > tiny -->

---

### if-else 

Example: What is the output of this code?
```{r, echo = TRUE, results = "hide"}

my_numbers <- c(3,4,5,"6",7)

if (!is.numeric(my_numbers)) {
  print("At least one element is not numeric. Only provide numbers.")
} else {
  print("All numeric elements")
}
```

<!--
```{r, echo = FALSE}
my_numbers <- c(3,4,5,"6",7)

if(!is.numeric(my_numbers)) {
  print("At least one element is not numeric. Only provide numbers.")
} else {
  print("All numeric elements")
}
```
-->

---

### if, else if, else

We can expand the basic **if-else** structure to incorporate **multiple tests** by adding **`else if`** statement(s) between the initial `if` and the final `else`


Syntax:
```
if (condition1) {
    # action performed when condition 1 is TRUE
    action1
} else if (condition2) {
    # action performed when condition 2 is TRUE
    action2
} else {
    # action performed when conditions 1 and 2 are FALSE
    action3
}
```
--

Conditional statements are evaluated **sequentially** (the order of your code matters!):

* conditions are checked one-by-one in the order they are written
* once R finds a TRUE condition, it stops, and the remaining conditions are not evaluated

---

### if, else if, else

Example: check if a given number is positive, negative, or zero
```{r, echo = TRUE, results = "hide"}

x <- 0

if (x > 0) {
  print("x is a positive number")
} else if (x < 0) {
  print("x is a negative number")
} else {
  print("x is zero")
}
```

---

### if, else if, else 

Another Example with multiple `else if`:

```{r, eval = FALSE, results = "hide"}

# take user input
temperature <- as.integer(readline(prompt = "Enter today's temperature in Celsius: "))

# determine weather based on temperature
if (temperature >= 30) {
  weather <- "Hot"
} else if (temperature >= 20) {
  weather <- "Cool"
} else if (temperature >= 10) {
  weather <- "Breezy"
} else {
  weather <- "Freezing!"
}

# print
print(weather)


```


---

### Nested if-else

Use nested if-else statements to specify **conditions within conditions**. What does this code do? 

<!-- The %% modulo operator calculates the remainder from a division operation. Here 15 : 2 = 7, 1 is the reminder, thus the number is odd -->

Example (the %% calculates the remainder from a division): 

.small[
```{r eval = FALSE, results = "hide"}
x <- 15

if (x > 0) {
  if (x %% 2 == 0) {
    print("x is a positive even number")
  } else {
    print("x is a positive odd number")
  }
} else {
  if (x %% 2 == 0) {
    print("x is a negative even number")
  } else {
    print("x is a negative odd number")
  }
}

```
]

---

### Nested if-else

Same example but check also the condition `x <- 0`:

```{r eval = FALSE, results = "hide"}
x <- 0

if (x > 0) {
  if (x %% 2 == 0) {
    print("x is a positive even number")
  } else {
    print("x is a positive odd number")
  }
} else if (x < 0) {
  if (x %% 2 == 0) {
    print("x is a negative even number")
  } else {
    print("x is a negative odd number")
  }
} else {
  print("x is zero")
}

```



---

### ifelse()

R accepts `if` and `else` statements, but also statements using the built-in **`ifelse()`**. They are similar but not the same and serve different purposes.  

.small[
Syntax:
```
ifelse (condition to be evaluated,
    action performed when condition is TRUE,
    action performed when condition is FALSE)
```

Example:
```{r}
y <- 3
ifelse(sqrt(16) > y, 
       sqrt(16),
       0)
```
] 

<!--
if-else: more flexibile and handles complex conditional logic; returns a scalar

ifelse(): simpler and more suited for simple assignments or to transform variables in a dataframe; returns a vector
-->

---

### ifelse()

Example: What is the output of this code?

```{r, echo = TRUE, results = "hide"}
numbers <- c(10, 6, 7)
ifelse(numbers %% 2 == 1, 
       "odd",
       "even")
```

--
```{r, echo = FALSE}
numbers <- c(10, 6, 7)
ifelse(numbers %% 2 == 1,
       "odd",  
       "even")
```

--

**Note the input here is a vector, not a single number, thus the output is also a vector**: the code will evaluate each element of the vector. 

This is because **`ifelse()` supports vectorized operations**: operations directly applied on entire vectors of data, rather than looping through individual elements one-by-one.

<!--
How can we rewrite the same code using the standard `if-else` statement?

```
# won't work
numbers <- c(10, 6, 7)
if (numbers %% 2 == 1) { 
  print("odd")
} else {
  print("even")
}
```

```
# need a loop
numbers <- c(10, 6, 7)
for (i in 1:length(numbers)) {
  if (numbers[i] %% 2 == 1) {
    print("odd")
    } 
else 
  print("even")
 }
```
-->

---

### if_else()

Also note, there is a `dplyr` version of the base R `ifelse()` function, called `if_else()`: they have the same logic, but the latter is better to use with the tidyverse

[documentation here](https://dplyr.tidyverse.org/reference/if_else.html)

Compare the two!

---

### ifelse()

Example: What is the output of this code?
```{r, echo = TRUE, results = "hide"}
qualify <- tibble("Athlet" = c("Noah", "Julio", "Nick", "Maria"), 
                      "Scores" = c(32, 37, 28, 30))

ifelse(qualify$"Scores" > 30, "Admitted", "Rejected")
```

--

Example: What is the output of this code?
```{r, echo = FALSE}
qualify <- data.frame("Athlet" = c("Noah", "Julio", "Nick", "Maria"), 
                      "Scores" = c(32, 37, 28, 30))

ifelse(qualify$"Scores" > 30, "Admitted", "Rejected")
```

---

### ifelse() 

Example: use `ifelse()` to recode variables in a dataframe...

The variable `decisionDirection` takes four values (1 conservative, 2 liberal, 3 unspecifiable, and NA). 

Recode it to take three values (0 conservative, 1 liberal, and NA for both NA and unspecifiable)

```{r eval = FALSE}
scdb_case %>%
  mutate(decisionDirection = ifelse(
    decisionDirection == 3, # condition to be evaluated
    NA,                     # if true, do NA
    decisionDirection - 1   # if false, subtract one to value
    ))
```

<!--
More on this:
https://docs.ycrc.yale.edu/r-novice-gapminder/07-control-flow/
Comparing ifelse() and for loop:
https://www.r-bloggers.com/2020/02/if-else-and-ifelse/
-->

---

class: inverse, middle

## Practice writing conditional statements

See today's class materials posted on the website

