---
title: "HW01: Practice editing .md and .Rmd"
date: 2022-10-02T13:30:00-06:00 
type: book
toc: true
draft: true
aliases: ["/hw01-edit-README.html"]
categories: []
weight: 10
---

<!--
title: "HW01: Practice editing .md and .Rmd"
date: 2022-10-02T13:30:00-06:00  # Schedule page publish date
#publishdate: 2019-03-01
draft: false
#type: post
aliases: ["/hw01-edit-README.html"]

summary: "Test software installation, GitHub setup, and homework submission process, as well as demonstrate basic competency in Markdown and R Markdown."
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Overview

**Due by 11:59 PM on Monday, October 9th.**

The goals of this first homework assignment are multiple (1) to test your software installation, your GitHub setup, and the homework submission process, and (2) to demonstrate competency in basic R syntax, Markdown and R Markdown. I strongly encourage you to submit this assignment earlier, so you are sure everything works as expected. 


# Accessing your `hw01` repository

* Go [at this link](ADD LINK) to accept and create your private `hw01` repository on GitHub. Once you do so, your repository will be built in a few seconds. It follows the naming convention `hw01-<USERNAME>`  
* Once the your repository has been created, click on the link you see, which will take you to your repository. 
* Finally, clone the repository to your computer (or R Workbench) following the process below.


# Cloning your `hw01` repository

After you have accessed the `hw01` repository (see above), follow these steps to clone it. Whenever possible, this will be the preferred route for setting up your R projects:

* In RStudio or RStudio Workbench, start a new Project with *File > New Project > Version Control > Git*. In the "repository URL" paste the URL of your newly created GitHub repository: go back to the repository, click on the green button "Code" and grab the correct link, either SSH or HTTPS (the one you used to set up your cache credentials, if you are using R Workbench this must be SSH). The repository name should automatically populate; if not, type your `hw01-<USERNAME>` as name. 
* Decide where to store the local directory for the project. Don't scatter everything around your computer - have a central location, or some meaningful structure. For example, for repositories you create in this course, you can setup a `css` directory and clone all your repos there.
* Before proceeding, check the little box "Open in new session", as that's what you'll usually do in real life.
* Click "Create Project" to create a new sub-directory, which will be all of these things:
    * a directory on your computer
    * a Git repository, linked to a remote GitHub repository
    * an RStudio Project

{{% callout note %}}
Before completing the above steps, ensure you followed the configuration steps [here](/setup/git-configure/), and everything works as expected.
{{% /callout %}}


# General workflow

Your general workflow for all homework assignments will be:

* Accept the repo and clone it (see above)
* Make changes locally to the files (for this assignment: `README.md` and `Rcode.rmd`) in RStudio
* Save your changes
* Pull-Stage-Commit-Push: 
  * Pull from GitHub (this is just an empty precaution now, e.g. nothing should happen, but it is part of the basic workflow and will matter when you collaborate with others)
  * Stage and Commit your changes to your local Git repo
  * Push them online to GitHub

You can complete these steps using the Git GUI integrated into RStudio -- if you have not yet, I strongly recommend to go through the [Using Git with R Studio](/setup/git/git-with-rstudio) tutorial to get comfortable with the workflow. Do not directly modify your online GitHub repo; instead modify your local Git repo, then stage-commit-push your changes up to your online GitHub repo. 


# Assigment descripion

All assignments for this course will be submitted using Markdown and R Markdown:
* [Markdown](https://daringfireball.net/projects/markdown/) (.md) is a lightweight and basic language that easily converts between file formats. For example, GitHub README files are frequently written in Markdown (that is they have an .md extension, or they are a text with extension .txt). Markdown files are directly rendered on the GitHub website (vs. R Markdown files which are not), and they are also well integrated into R Markdown. GitHub includes a [guide](https://guides.github.com/features/mastering-markdown/) for writing Markdown documents.
* [R Markdown](https://rmarkdown.rstudio.com/) (.rmd or .Rmd) is a fancy version of Markdown, designed for R. It combines R code, output, and written text into a single document. We will use R Markdown a lot for homework assignments.

For this assignment, you need to modify and push to your GitHub homework repository the following two files: `README.md` and `Rcode.rmd`


## `README.md` 

You will notice there is already a `README.md` file in the repo. A README is usually a plain .txt or .md file. The purpose of a README file in a repository is to communicate important information about your project (software, how to run the code, preview of the results, etc.). The goal of this assignment is to practice generating and editing a README file.

Your task is to edit this file by adding a brief biography of yourself. To achieve full marks, your README should include the following elements:

* Headers (one or more)
* Emphasis (italics or/and bold)
* At least one list
* Images: add a picture (of yourself or something else) to your repo and embed it in your README
* At least one link
* A detailed summary of the Git/GitHub workflow you adopted for this homework
* An evaluation of your experience with this homework (e.g., write about something new you learned, something that surprised you, etc.); these will be a requirement for every homework and they should always be placed in the readme file

## `Rcode.rmd`

You will notice there is already a `Rcode.rmd` file in the repo. Your task is to edit this file by adding more YAML arguments (see readings) and at least two R code chunks. To achieve full marks, your `Rcode.rmd` should include the following elements:

* At least one new YAML Header argument. You can modify one of three arguments that are already there or add a new one to them (e.g., author, subtitle, table of contents, etc.)
* At least two R code chunks:
  * You are welcome to rely on the in-class materials as starting point, but do not copy the exact same code: make changes to customize or expand it
  * All code chunks must execute correctly and your final knitted document (which you should submit as well, as md or html) must display both the code and the results
  * Before each code chunk, add a brief text that explains what each code chunk does


# Submit the assignment

To submit the assignment, follow these steps:
* First push to your repository the last version of your assignment before the deadline. You GitHub repository should contain: the modified `README.md` and `Rcode.rmd`, plus the final knitted document that you generated starting from the `Rcode.rmd`(this document should be an .md or html file)
* When you are ready to submit, copy your repository URL (e.g. https://github.com/css-fall23/hw1-brinasab) and submit it to Canvas under HW01 before the deadline. Do not submit files on Canvas, we only need the link to your directory




# Rubric

Needs improvement: `README.md` says the equivalent of "This is the repository of Name Surname" with little details. `Rcode.rmd` contains R code chunks, but no explanation nor additional YAML headers are added and/or the final knitted document has not been submitted. All work done via Git/Github ... but that's just a guess, because the student doesn't explain in detail how it was done.

Good: Something in between.

Excellent: `README.md` provides a proper introduction of the student. It also demonstrates experimentation with several aspects of the Markdown syntax (e.g., section headers, links, bold, italic, bullet points, image, etc.). The student provides a detailed description on how they got the changes into `README.md` and offers a few reflections on their Git/GitHub workflow and their experience with Markdown. `Rcode.rmd` knits properly and contains one or more new YAML headers, and novel R code chunks with a detailed explanation of them. All required files have been submitted.

For further details, see the [general rubric](/faq/homework-evaluations/) we adopt for grading.


# Acknowledgments

```{r child = here::here("R", c("_ack_stat545.Rmd", "_ack_ben.Rmd"))}
```
