---
title: "Strings & Regular Expressions"
author: "MACSS 30500 <br /> University of Chicago"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: r
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
# generate CSS file
library(xaringanthemer)
style_duo_accent(
  primary_color = "#011f4b",
  secondary_color = "#bbd6c7",
  inverse_header_color = "#222222",
  black_color = "#222222",
  header_font_google = xaringanthemer::google_font("Atkinson Hyperlegible"),
  text_font_google = xaringanthemer::google_font("Atkinson Hyperlegible"),
  code_font_google = xaringanthemer::google_font("Source Code Pro"),
  base_font_size = "24px",
  code_font_size = "20px",
  # title_slide_background_image = "https://github.com/uc-dataviz/course-notes/raw/main/images/hexsticker.svg",
  # title_slide_background_size = "contain",
  # title_slide_background_position = "top",
  header_h1_font_size = "2rem",
  header_h2_font_size = "1.75rem",
  header_h3_font_size = "1.5rem",
  extra_css = list(
    "h1" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    "h2" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    "h3" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    ".tiny" = list("font-size" = "70%"),
    ".small" = list("font-size" = "90%"),
    ".midi" = list("font-size" = "150%"),
    ".tiny .remark-code" = list("font-size" = "70%"),
    ".small .remark-code" = list("font-size" = "90%"),
    ".midi .remark-code" = list("font-size" = "150%"),
    ".large" = list("font-size" = "200%"),
    ".xlarge" = list("font-size" = "600%"),
    ".huge" = list(
      "font-size" = "400%",
      "font-family" = "'Montserrat', sans-serif",
      "font-weight" = "bold"
    ),
    ".hand" = list(
      "font-family" = "'Gochi Hand', cursive",
      "font-size" = "125%"
    ),
    ".task" = list(
      "padding-right" = "10px",
      "padding-left" = "10px",
      "padding-top" = "3px",
      "padding-bottom" = "3px",
      "margin-bottom" = "6px",
      "margin-top" = "6px",
      "border-left" = "solid 5px #F1DE67",
      "background-color" = "#F3D03E"
    ),
    ".pull-left" = list(
      "width" = "49%",
      "float" = "left"
    ),
    ".pull-right" = list(
      "width" = "49%",
      "float" = "right"
    ),
    ".pull-left-wide" = list(
      "width" = "70%",
      "float" = "left"
    ),
    ".pull-right-narrow" = list(
      "width" = "27%",
      "float" = "right"
    ),
    ".pull-left-narrow" = list(
      "width" = "27%",
      "float" = "left"
    ),
    ".pull-right-wide" = list(
      "width" = "70%",
      "float" = "right"
    ),
    ".blue" = list(color = "#2A9BB7"),
    ".purple" = list(color = "#a493ba"),
    ".yellow" = list(color = "#f1de67"),
    ".gray" = list(color = "#222222")
  )
)

source(here::here("R", "slide-opts.R"))
xaringanExtra::use_panelset()
```

```{r pkgs, include = FALSE, cache = FALSE}
options(htmltools.dir.version = FALSE)

library(tidyverse)
library(tidytext)
library(scales)
library(here)
library(patchwork)
library(magrittr)
library(countdown)

set.seed(1234)
theme_set(theme_minimal(base_size = 16))
```

class: inverse, middle

# Agenda

* Strings: definition
* Regular Expressions: definition and uses
* The `stringr()` package
* Examples and practice (today's class materials)

<!-- we focus mostly on example and practice with Rmd file from in-class materials! -->



---

class: inverse, middle

## Strings


---

### Strings

```{r}
string1 <- "This is a string"
string2 <- 'If I want to include a "quote" inside a string, I use single quotes'
```

--

**Today we talk about strings and methods for manipulating strings using regular expressions!**

<!-- ask: how many of you have heard or had the pleasure of working with reg expr?** -->

--

This lecture complements today's readings (check them out, especially the ["Extracting data from Strings"](https://r4ds.hadley.nz/strings#extracting-data-from-strings)). Today, we'll dive deeper into regular expressions than the book does.


<!-- Not 1:1 allignment but haver focus on Reg Expr because super improtant to know them! 
So formally a string is... -->

---

### String and Character Vector

A **string** is a sequence of characters (a piece of text) enclosed in either double `"` or single quotes `'`. Strings can include letters, numbers, symbols, and whitespace characters.

```{r}
string <- "Ciao, my name is Sabrina and I am 99 years old!!"
class(string)
length(string)
```

---

### String and Character Vector

A **character vector** is a collection of strings: each element of the vector is a string! So, a single string is essentially a character vector of length one.

```{r}
char_vect <- c("Ciao come stai?", "Ciao", "Hello", "etc.") 
class(char_vect)
length(char_vect)
```

We use character vectors when we need to handle multiple pieces of text together (e.g., when we handle multiple strings)!

---

class: inverse, middle

## Regular Expressions


---

### Regular Expressions: definitions and uses

Regular expressions ("regex" or "regexes") **are strings containing normal characters and/or special meta-characters.** They describe a specific pattern to match in a given text.

More formally, regex is:
* a *language on its own right* 
* used for *pattern matching* 
* in many programming languages, such as R, Python, and others!

--

Our goal today is to decipher and write patterns like this:

```{r}
email_pattern <- "[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\\.[a-zA-Z]{2,}"
```


<!--
But then you might ask yourself...
given our ability to manipulate strings and our ability to test for equivalence (with the == operator) 
or to test whether some string contains another (with the %in% operator), we don't technically need regular expressions... Not exactly! We want to use regex becasue they are POWERFUL
-->

---


### Regular Expressions: definitions and uses

Regular expressions are powerful!

They allow us to find patterns **in any task that deals with text,** such as NLP or data-cleaning tasks. For example, you can use regular expressions to:

* Extract specific elements from texts (e.g. dates, find words that include a given set of letters, find all past tenses in a text,)
* Perform textual substitutions (e.g. find and replace HTML tags after you scraped a page)
* Verify text format (e.g. "Is this email/phone-number/address valid?")

`stringr` and regular expressions will be particularly useful for those of you who plan to work with textual data in this course and beyond.  


---

class: inverse, middle

## Regular Expression with `stringr()`

---

### The `stringr()` package in R

When you use regular expressions for your analysis, most likely you will need to use your regular expression together with one of the functions from the `stringr()` package. 

This package includes several functions that let you: detect matches in a string, count the number of matches, extract them, replace them with other values, split a string based on a match, etc. 

---

### The `stringr()` package in R

Fundamental `stringr()` functions and their use:

* `str_view()` return the first regex match
* `str_view_all()` return all regex matches (deprecated in the last version of `stringr`)
* `str_detect()`: detect matches in a string
* `str_count()`: count the number of matches
* `str_extract()` and `str_extract_all()`: extract matches
* `str_replace()` and `str_replace_all()`: replace matches
* `str_split()`: split a string based on a match

---

### Resources

* [`stringr()` documentation and cheatsheet](https://stringr.tidyverse.org/)
* [Chapter 15](https://plsc-31101.github.io/course/strings-and-regular-expressions.html#applying-regex) by Rochelle Terman, explains `stringr()` (using the *R for Data Science* textbook)  
* Read [Chapter 14 "Strings"](https://r4ds.hadley.nz/strings) in "R for Data Science" 2nd Edition (read it all, but especially 14.4 "Extracting data from strings").
* Read [Chapter 17](https://bookdown.org/rdpeng/rprogdatascience/regular-expressions.html#the-stringr-package) from *R Programming for Data Science*. This book covers the entire range of regular expressions packages and functions: you do not need to understand everything, focus on the big picture. In-class we talk about `stringr()`
* For more in-depth info on regular expressions check [this excellent tutorial](https://github.com/ziishaned/learn-regex/blob/master/README.md)

---

class: inverse, middle

## Examples and Practice

Download today's class material from the website to follow along

