---
title: "More dpyr and data cleaning"
author: "MACSS 30500 <br /> University of Chicago"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: r
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
# generate CSS file
library(xaringanthemer)
style_duo_accent(
  primary_color = "#011f4b",
  secondary_color = "#bbd6c7",
  inverse_header_color = "#222222",
  black_color = "#222222",
  header_font_google = xaringanthemer::google_font("Atkinson Hyperlegible"),
  text_font_google = xaringanthemer::google_font("Atkinson Hyperlegible"),
  code_font_google = xaringanthemer::google_font("Source Code Pro"),
  base_font_size = "26px",
  code_font_size = "20px",
  # title_slide_background_image = "https://github.com/uc-dataviz/course-notes/raw/main/images/hexsticker.svg",
  # title_slide_background_size = "contain",
  # title_slide_background_position = "top",
  header_h1_font_size = "2rem",
  header_h2_font_size = "1.75rem",
  header_h3_font_size = "1.5rem",
  extra_css = list(
    "h1" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    "h2" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    "h3" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    ".tiny" = list("font-size" = "70%"),
    ".small" = list("font-size" = "90%"),
    ".midi" = list("font-size" = "150%"),
    ".tiny .remark-code" = list("font-size" = "70%"),
    ".small .remark-code" = list("font-size" = "90%"),
    ".midi .remark-code" = list("font-size" = "150%"),
    ".large" = list("font-size" = "200%"),
    ".xlarge" = list("font-size" = "600%"),
    ".huge" = list(
      "font-size" = "400%",
      "font-family" = "'Montserrat', sans-serif",
      "font-weight" = "bold"
    ),
    ".hand" = list(
      "font-family" = "'Gochi Hand', cursive",
      "font-size" = "125%"
    ),
    ".task" = list(
      "padding-right" = "10px",
      "padding-left" = "10px",
      "padding-top" = "3px",
      "padding-bottom" = "3px",
      "margin-bottom" = "6px",
      "margin-top" = "6px",
      "border-left" = "solid 5px #F1DE67",
      "background-color" = "#F3D03E"
    ),
    ".pull-left" = list(
      "width" = "49%",
      "float" = "left"
    ),
    ".pull-right" = list(
      "width" = "49%",
      "float" = "right"
    ),
    ".pull-left-wide" = list(
      "width" = "70%",
      "float" = "left"
    ),
    ".pull-right-narrow" = list(
      "width" = "27%",
      "float" = "right"
    ),
    ".pull-left-narrow" = list(
      "width" = "27%",
      "float" = "left"
    ),
    ".pull-right-wide" = list(
      "width" = "70%",
      "float" = "right"
    ),
    ".blue" = list(color = "#2A9BB7"),
    ".purple" = list(color = "#a493ba"),
    ".yellow" = list(color = "#f1de67"),
    ".gray" = list(color = "#222222")
  )
)

source(here::here("R", "slide-opts.R"))
knitr::opts_chunk$set(echo = FALSE)
```

```{r pkgs, include = FALSE, cache = FALSE}
library(tidyverse)
library(here)
library(knitr)
library(countdown)
#library(tidyexplain)
library(rcis)

theme_set(theme_minimal(base_size = rcis::base_size))
```

```{r anim-opts, include = FALSE}
#set_anim_options(anim_opts = anim_options(text_size = 4, cell_width = 2, title_size = 14))
```

class: inverse, middle

# Agenda

Review and expand our knowledge of:
* factors 
* missing data
* `dplyr` for data transformation
 

---

class: inverse, middle

## Review of factors 

---

Pull from last lecture (lecture 4) slides

---

class: inverse, middle

## Practice using factors 

Download today's in-class materials and complete the exercises on working with factors.

---

class: inverse, middle

## Missing data

We talk about missing data using the `missing-data.Rmd` document, available in today's in-class materials on the website.

---

class: inverse, middle

## `dplyr` verbs for data transformation

Today we will review the verbs we already know with additional examples and introduce a few new ones.

You can follow along by using the `more-dplyr.Rmd` document, available in today's in-class materials on the website.

---

<!-- last week... -->

### Flow for data transformation

Conceptually, we have learned that to perform data transformations we need to:
1. Get the data frame
1. Use `dplyr` verbs/functions to tell R what to do with the data frame
1. Generate a new data frame with the result 


Today we review the main verbs we already know (see next slide) with more examples, and we learn new ones

---

These are the **main verbs** you will use to transform your data. By **combining them together**, you can perform powerful data manipulation tasks.

`function()`  | Action performed
--------------|--------------------------------------------------------
`filter()`    | Picks observations from the data frame based on their values (operates on rows)
`arrange()`   | Changes the order of observations based on their values
`select()`    | Picks variables from the data frame based on their names (operates on columns)
`rename()`    | Changes the name of columns in the data frame
`mutate()`    | Creates new columns from existing ones
`group_by()`  | Changes the unit of analysis from the complete data frame to individual groups
`summarize()` | Collapses the data frame to a smaller number of rows to summarize the larger data
