---
title: "Functions in R"
author: "MACSS 30500 <br /> University of Chicago"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: r
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
# generate CSS file
library(xaringanthemer)
style_duo_accent(
  primary_color = "#011f4b",
  secondary_color = "#bbd6c7",
  inverse_header_color = "#222222",
  black_color = "#222222",
  header_font_google = xaringanthemer::google_font("Atkinson Hyperlegible"),
  text_font_google = xaringanthemer::google_font("Atkinson Hyperlegible"),
  code_font_google = xaringanthemer::google_font("Source Code Pro"),
  base_font_size = "24px",
  code_font_size = "20px",
  # title_slide_background_image = "https://github.com/uc-dataviz/course-notes/raw/main/images/hexsticker.svg",
  # title_slide_background_size = "contain",
  # title_slide_background_position = "top",
  header_h1_font_size = "2rem",
  header_h2_font_size = "1.75rem",
  header_h3_font_size = "1.5rem",
  extra_css = list(
    "h1" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    "h2" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    "h3" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    ".tiny" = list("font-size" = "70%"),
    ".small" = list("font-size" = "90%"),
    ".midi" = list("font-size" = "150%"),
    ".tiny .remark-code" = list("font-size" = "70%"),
    ".small .remark-code" = list("font-size" = "90%"),
    ".midi .remark-code" = list("font-size" = "150%"),
    ".large" = list("font-size" = "200%"),
    ".xlarge" = list("font-size" = "600%"),
    ".huge" = list(
      "font-size" = "400%",
      "font-family" = "'Montserrat', sans-serif",
      "font-weight" = "bold"
    ),
    ".hand" = list(
      "font-family" = "'Gochi Hand', cursive",
      "font-size" = "125%"
    ),
    ".task" = list(
      "padding-right" = "10px",
      "padding-left" = "10px",
      "padding-top" = "3px",
      "padding-bottom" = "3px",
      "margin-bottom" = "6px",
      "margin-top" = "6px",
      "border-left" = "solid 5px #F1DE67",
      "background-color" = "#F3D03E"
    ),
    ".pull-left" = list(
      "width" = "49%",
      "float" = "left"
    ),
    ".pull-right" = list(
      "width" = "49%",
      "float" = "right"
    ),
    ".pull-left-wide" = list(
      "width" = "70%",
      "float" = "left"
    ),
    ".pull-right-narrow" = list(
      "width" = "27%",
      "float" = "right"
    ),
    ".pull-left-narrow" = list(
      "width" = "27%",
      "float" = "left"
    ),
    ".pull-right-wide" = list(
      "width" = "70%",
      "float" = "right"
    ),
    ".blue" = list(color = "#2A9BB7"),
    ".purple" = list(color = "#a493ba"),
    ".yellow" = list(color = "#f1de67"),
    ".gray" = list(color = "#222222")
  )
)

source(here::here("R", "slide-opts.R"))
```

```{r pkgs, include = FALSE, cache = FALSE}
library(tidyverse)
library(palmerpenguins)
library(here)
library(countdown)

set.seed(1234)
theme_set(theme_minimal(base_size = rcis::base_size))
```

class: inverse, middle

# Functions

---

### Using existing functions

Functions are all over R! We have been **using functions** from day 1 of this class. For example:
```{r}
a <- c(1:10)
a
mean(a)
sqrt(a)
```

---

### Using existing functions

To use a function, we need to know its arguments and pass them correctly. Type `help(function name)` in the console to learn more about a function behavior and its arguments:
```{r}
help(mean)
help(sqrt)
```

--

Some functions come from base R, like those I listed here, other functions have been developed from packages (e.g., all the functions from the tidyverse that we have been using). 

---

### Writing your own function

Other than using already defined functions, we can **create our own function**! 

A function has the following components:

* name and argument(s)
* body
* return statement

Syntax:
```
name <- function(arg1, arg2, ...) {  # name of function and arguments (inputs) it takes
    value <- body_of_function        # the body is the code that manipulates arguments
    return(value)                    # what the function returns (results)
    }
```
<!-- we can assign the function to a named object like any other object -->

---

### Writing your own function

For example, let's imagine the built-in `mean()` function would not exist... we could **write our own mean function** to calculate the mean of a vector of numbers.

```
name <- function(arg1, arg2, ...) {  # name of function and arguments (inputs) it takes
    value <- body_of_function        # the body is the code that manipulates arguments
    return(value)                    # what the function returns (results)
    }
```

``` {r}
my_mean <- function(vector) {
    tot_values <- length(vector)
    m <- (sum(vector/tot_values))
    return(m)
}
```

``` {r}
# short version
my_mean <- function(vector) {
  sum(vector/length(vector))
}
```

---

### Writing your own function

Notice, like any other piece of code, you can write a function by specifying all the intermediate steps in the body, or you can "skip" passages. Here are two different ways to write the same function: the first is easier to read, the second is more compact.

``` {r}
# long version
my_mean <- function(vector) {
    tot_values <- length(vector)
    m <- (sum(vector/tot_values))
    return(m)
}
```

``` {r}
# short version
my_mean <- function(vector) {
  return(sum(vector/length(vector)))
}
```

---

### Writing your own function

To make a function work, we need to 'call' it with some values:
``` {r}
my_mean <- function(vector) {
  return(sum(vector/length(vector)))
}
```

```{r}
a <- c(1:10)
my_mean(vector = a)
b <- c(1, 3, 6, 9, 20, 30)
my_mean(b)
```
This function has one argument: `vector`. When we call the function, this argument will take a concrete vector, such as `a`, `b`, etc. Notice, we can, but we do not have to, specify the argument name.

---

### Same code without a function

Without using a function, we have to type the code to calculate the mean every time we need it:
```{r}
# calculate the mean of vector a
a <- c(1:10)
n <- length(a)
m <- (sum(a/n))
m
```

```{r}
# calculate the mean of vector b
b <- c(1, 3, 6, 9, 20, 30)
n <- length(b)
m <- (sum(b/n))
m
```

This code is redundant, more prone to errors, and less organized.

<!--
```{r}
b <- c(1, 3, 6, 9, 20, 30)
m <- (sum(b/length(b)))
m
```
-->

---

### So... What is a function?

A function is **code that is organized and named so that it can be executed by simply executing the name** (referred to as ‘calling’ a function). 

Functions:
* take some values as input and return output values
* reduce repetitive code
* make it easier to reuse code
* organize code (e.g., one function imports the data, another cleans the data, another split text into words, etc.)

 

---

### Writing your own function: things to notice

Names:

* provide **unique names for your functions** (e.g., `my_mean` vs `mean`); if you accidentally choose a function name that matches the name of existing R function, your function will override the existing function and only your function will be active in your current session

*  unless you have a reason to do so, **do not create a new function if one already exists** (e.g., you do not want to create a `my_mean` function since `mean` is already defined in base R)

* provide **informative names** to your function: the name should tell you what the function does 

---

### Writing your own function: things to notice

Variables defined inside the function are **not available outside the function**:

``` {r}
my_mean <- function(vector) {
    tot_values <- length(vector)
    mean_var <- (sum(vector/tot_values))
    return(mean_var)
}
```

```{r, error = TRUE}
tot_values
```

```{r, error = TRUE}
mean_var
```

---

### Writing your own function: things to notice

If you do not write a **`return` statement**, the last statement will be your output:
```{r}
my_mean <- function(vector) {
    tot_values <- length(vector)
    return(sum(vector/tot_values))
}

my_mean(c(1:10))
```

However, it is good practice to write a `return` statement. Everything piece of code after your `return` statement will be ignored. 

---

### stop()

Let’s define a function `fahr_to_kelvin()` that converts temperatures from Fahrenheit to Kelvin. Example taken from: https://swcarpentry.github.io/r-novice-gapminder/10-functions/index.html

```{r}
fahr_to_kelvin <- function(temp) {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}
```

```{r}
# freezing point of water
fahr_to_kelvin(32)
```

---

### stop()

For the `fahr_to_kelvin` function to work as intended, the argument `temp` must be a numeric value. Otherwise, the conversion between the two temperature scales will not work. 

To create an error, we can use the function `stop()`: we can use an if statement to check that the argument `temp` is a numeric value, and throw and error if that's not the case:

```{r}
fahr_to_kelvin <- function(temp) {
  if (!is.numeric(temp)) {
    stop("temp must be a numeric vector.")
  } else {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin) 
  }
}
```

```{r, error = TRUE}
# freezing point of water
fahr_to_kelvin("32")
```

---

class: inverse, middle

# Practice writing functions
