---
title: "Functions"
author: "MACSS 30500 <br /> University of Chicago"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: r
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
# generate CSS file
library(xaringanthemer)
style_duo_accent(
  primary_color = "#011f4b",
  secondary_color = "#bbd6c7",
  inverse_header_color = "#222222",
  black_color = "#222222",
  header_font_google = xaringanthemer::google_font("Atkinson Hyperlegible"),
  text_font_google = xaringanthemer::google_font("Atkinson Hyperlegible"),
  code_font_google = xaringanthemer::google_font("Source Code Pro"),
  base_font_size = "24px",
  code_font_size = "20px",
  # title_slide_background_image = "https://github.com/uc-dataviz/course-notes/raw/main/images/hexsticker.svg",
  # title_slide_background_size = "contain",
  # title_slide_background_position = "top",
  header_h1_font_size = "2rem",
  header_h2_font_size = "1.75rem",
  header_h3_font_size = "1.5rem",
  extra_css = list(
    "h1" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    "h2" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    "h3" = list(
      "margin-block-start" = "0.4rem",
      "margin-block-end" = "0.4rem"
    ),
    ".tiny" = list("font-size" = "70%"),
    ".small" = list("font-size" = "90%"),
    ".midi" = list("font-size" = "150%"),
    ".tiny .remark-code" = list("font-size" = "70%"),
    ".small .remark-code" = list("font-size" = "90%"),
    ".midi .remark-code" = list("font-size" = "150%"),
    ".large" = list("font-size" = "200%"),
    ".xlarge" = list("font-size" = "600%"),
    ".huge" = list(
      "font-size" = "400%",
      "font-family" = "'Montserrat', sans-serif",
      "font-weight" = "bold"
    ),
    ".hand" = list(
      "font-family" = "'Gochi Hand', cursive",
      "font-size" = "125%"
    ),
    ".task" = list(
      "padding-right" = "10px",
      "padding-left" = "10px",
      "padding-top" = "3px",
      "padding-bottom" = "3px",
      "margin-bottom" = "6px",
      "margin-top" = "6px",
      "border-left" = "solid 5px #F1DE67",
      "background-color" = "#F3D03E"
    ),
    ".pull-left" = list(
      "width" = "49%",
      "float" = "left"
    ),
    ".pull-right" = list(
      "width" = "49%",
      "float" = "right"
    ),
    ".pull-left-wide" = list(
      "width" = "70%",
      "float" = "left"
    ),
    ".pull-right-narrow" = list(
      "width" = "27%",
      "float" = "right"
    ),
    ".pull-left-narrow" = list(
      "width" = "27%",
      "float" = "left"
    ),
    ".pull-right-wide" = list(
      "width" = "70%",
      "float" = "right"
    ),
    ".blue" = list(color = "#2A9BB7"),
    ".purple" = list(color = "#a493ba"),
    ".yellow" = list(color = "#f1de67"),
    ".gray" = list(color = "#222222")
  )
)

source(here::here("R", "slide-opts.R"))
```

```{r pkgs, include = FALSE, cache = FALSE}
library(tidyverse)
library(palmerpenguins)
#library(here)
#library(countdown)

set.seed(1234)
theme_set(theme_minimal(base_size = rcis::base_size))
```

class: inverse, middle

## Agenda

**From using to writing functions:**
  * Using existing functions
  * Writing your own function
  * Practice
  
**More advanced content**
  * The return statement 
  * Use `stop()`
  * Anonymous functions
  * Practice 
  * Writing more complex functions (example)
  * Functions with more than one argument
  * Functions as a way to organize your code
  * Documenting your function
  * Writing functions in the `tidyverse`

<!-- maybe we stop after the second practice! I improved and enlarged this lecture from last yer and not sure how long it ll take AND do not want to overwhelm you
So we ll check in after the second practice exercises and see how things go 
we have time built in to finish this tomorrow! --> 

---

class: inverse, middle

## From using to writing functions

  * Using existing functions
  * Writing your own function
  * Practice

---

### Using existing functions

Functions are all over R! We have been **using functions** from day 1 of this class. For example:
```{r}
a <- c(1:5)
a
sum(a)
mean(a)
```


---

### Using existing functions
 
A function: 

* is a block of code that accomplishes a task
* has a name that typically describes the task the function performs, like `sum()` or `mean()`
* takes one or more inputs (e.g., arguments), processes it, and returns one or more outputs

---

### Using existing functions

To use a function, we need to know its arguments and pass them correctly. 

In the console, type `help()` with the function's name in parenthesis to learn more about a function behavior and its arguments:
```{r}
help(sum)
help(mean)
```

--

Some functions come from base R, like those above.

Other functions belong to packages, like all the functions we have been using, like `summarize()` or `mutate()` from `dplyr`

---

### Writing functions

Other than **using** functions written by someone else, we can **create our own function**! When we write our own function we need to provide the following:

* **name**: typically something that describes the task that the function performs
* **arguments (input)**: at least one data structure (e.g., value, vector, etc.)
* **body**: self-contained lines of code that manipulate the input, defining what the function does
* **return (output):** statement that returns the result of the function's calculations

Syntax:
```
name <- function(arg1, arg2, ...) { 
    result <- body with code that does things using arg1, arg2, etc.
    return(result)
    }
```

<!--
Notice: 
* we can assign the function to a named object like any other object
* typically the body is multiple lines with multiple intermediate variables
-->

---

### Writing functions

For example, rather than using the built-in `sum()` function, we could **write our own sum function** to calculate the sum of the elements of a numeric vector.

Syntax:
```
name <- function(arg1, arg2, ...) { 
    result <- body with code that does things using arg1, arg2, etc.
    return(result)
    }
```

Our sum function: 
```{r}
my_sum <- function(vector) {
  total <- 0
  for (element in vector) {
    total <- total + element
  }
  return(total)
}
```

<!-- LET'S STOP A SECOND AN UNPACK THIS CODE! DO THIS ON WORKBENCH -->

```{r, eval = FALSE, echo = FALSE}
# do not evaluate, do not display

# define a vector
x <- c(2,1,5,2)

# our goal
sum(x)

# the math
sum_x = 2 + 1 + 5 + 2
#mean_x = (1 + 2 + 3 + 4) / 4 

# how we do this programmatically with code for this specific vector b?
total <- 0
for (element in x) {
  #print(element)
  total <- total + element
  #print(total)
}
total

# now that we know it works let's put it into a function 
# and test it out with other vectors
my_sum <- function(vector) {
  total <- 0
  for (element in vector) {
    total <- total + element
  }
  return(total)
}

# test with vector x
my_sum(x)

# test with another vector 
y <- c(4,3,7)
my_sum(y)
```

---

### Testing our sum function

To use a function, we need to **call it** with specific values. Our custom function `my_sum()` takes one argument, `vector`. When we call the function, this will be a concrete vector, such as `x`, `y`, etc. 

```{r}
x <- c(2,1,5,2)
my_sum(vector = x)  # you can specify the argument name but it's optional
```

```{r}
y <- c(4,3,7)
my_sum(y)
```

<!-- SKIP 
````
#j <- c("1", 6, 6)
#my_sum(j)

my_sum <- function(vector) {
  
  # convert vector to numeric, if possible
  # R coerces numeric strings like "1" to numeric values 1 (try if this was "one" )
  vector <- as.numeric(vector)
  #print(vector)
  #print(class(vector))
  
  # same code as above
  total <- 0
  for (v in vector) {
    total <- total + v
  }
  return(total)
}
```
-->

---

### Advantages of using functions

Use functions to...

* reduce repetitive code and chances for mistakes
* reuse code
* organize code (e.g., one function imports the data, another cleans the data, etc.)


<!--
Can do with a loop
```
v <- list(a, b)
output <- vector(mode = "list", length = length(v))

for (i in seq_along(v)) {
  output[[i]] <- (sum(v[[i]]/length(v[[i]])))
  #output[[i]] <- (mean(v[[i]]))
}
output
```
-->

---

### Things to remember when writing functions: name

Use **unique names** 
* if your function name matches the name of an existing R function, your function will replace the existing function for your current session (e.g., write `my_sum`, do not re-write `sum`)
* unless there is a specific reason (e.g. learning purposes, etc.), refrain from creating a new function like `my_sum()` when R already has a built-in `sum()` function you can use

Use **informative names**: 
* the name should tell you what the function does 
* it should not be too long
* avoid reserved words like `if`, `else`, `for`, `function`, etc. To see the full list of reserved words type `help(reserved)` in the console

---

### Things to remember when writing functions: variables scope

Variables defined inside a function are **not available outside it.** Their scope lies within and is limited to the function itself:

```{r, error = TRUE}
my_sum <- function(input_vector) {
  total_elements <- 0
  for (element in input_vector) {
    total_elements <- total_elements + element
  }
  return(total_elements)
}

my_sum(c(1,2,3))
total_elements
```

---

class: inverse, middle

## Practice writing functions (part 1)

Download today's class materials from the website. Complete questions 1 and 2.

---

class: inverse, middle

## More about functions!

  * The return statement 
  * Use `stop()`
  * Anonymous functions
  * Practice 
  * Writing more complex functions (example)
  * Functions with more than one argument
  * Functions as a way to organize your code
  * Documenting your function
  * Writing functions in the `tidyverse`

<!-- maybe stop after practice? --> 

---

### The role of the return statement(s)

If you do not write a `return` statement, your output will be the last statement in your code:

```{r}
my_sum <- function(input_vector) {
  total_elements <- 0
  for (element in input_vector) {
    total_elements <- total_elements + element
  }
  total_elements
}

my_sum(c(1,2,3))
```

--

Tips:
* Use `return` statements to enhance code clarity and readability
* Any code after your `return` statement will be ignored

---

### Write multiple return statements

Use multiple returns using `if-else`:

```{r}
check_number <- function(x) {
  if (x > 0) {
    return("positive")
    }
  else if (x < 0) {
    return("negative")
    }
  else {
    return("zero")
    }
}
check_number(1)
```

If x > 0, the function returns "positive" without evaluating rest of the body.

---

### Write one return statement but with multiple outputs

Return multiple objects and collect them into a list or a vector:

```{r}
my_sum <- function(input_vector) {
  total_elements <- 0
  for (element in input_vector) {
    total_elements <- total_elements + element
  }
  return(list(total_elements, element)) # why is this a list? try without!
}

my_sum(c(1,2,3))
```

<!--

```
my_mean <- function(vector) {
    total_values <- length(vector)
    result <- (sum(vector/total_values))
    return(list(total_values, result))
}
my_mean(1:10)
```
-->

---

### Using `stop()` when writing a function

Define a function `celsius_to_fahr()` that converts temperatures from Celsius to Fahrenheit using the formula `fahrenheit = (celsius * 9/5) + 32`

```{r}
celsius_to_fahr <- function(celsius) {
  fahr <- (celsius * 9/5) + 32
  return(fahr)
}
```

```{r}
celsius_to_fahr(0)
celsius_to_fahr(-20)
```

---

### Using `stop()` when writing a function

To ensure this function works properly, the argument must be numeric. Otherwise, the conversion between the two temperature scales will not work.

Solution: use `if-else` to verify if the provided argument is numeric; if not, use `stop()` to raise an error and stop the function execution

```{r}
celsius_to_fahr <- function(celsius) {
  if (!is.numeric(celsius)) {
    stop("input must be a number")
  } else {
    fahr <- (celsius * 9/5) + 32
    return(fahr) 
  }
}
```

```{r, error = TRUE}
celsius_to_fahr("zero")
```

<!-- Why use `stop()` and not `print()`?
`stop():` interrupts the execution of the function and throw an error message when a condition is not met 
`print()`: doesn not stop the function execution just print the message
so the first enforces a condition that must be met for the function to execute, and prints a message (which is optional); the other is only a message
-->

---

### Anonymous functions 

<!-- You won't define anonymous or unnamed functions a lot, but you need to know they exist and recognize their syntax.-->

Anonymous functions are **functions without a name**.

When to use anonymous functions?
* when the function is short and you need it only once 
* when you use it in conjunction with other functions, such as those from the `apply()` family

---

### Anonymous functions

Imagine we have the following function:
```{r}
add <- function(x) {
  x + 3
}
add(2)
```

--

Re-write it as an anonymous function. Note: one line, no name, and call it with `()`:
```{r}
(function(x) {x + 3}) (2)
```

--

Often anonymous functions are written without the `{}`, like this:
```{r}
(function(x) x + 3)(2)
```

<!--
Here's an unammed function for calculating the mean of a vector `x`. In the following example, the input `x` to the function is each element of the list `l`.
```{r}
l <- list(1:5, 5:7)
lapply(l, FUN = function(x){sum(x)/length(x)})
```

see http://adv-r.had.co.nz/Functional-programming.html#anonymous-functions
-->

---

### Anonymous functions: with `sapply()` and `lapply()`

Same function as before but we pass a vector rather than a single number:
```{r}
add <- function(x) {
  x + 1
}
add(c(1,2,3))

```

To rewrite it as anonymous function, this code won't work:
```{r, error = TRUE}
(function(x) x + 1) (1,2,3)
```

It only works on single values, not separated by commas:
```{r}
(function(x) x + 1) (1:3)
```

---

### Anonymous functions: with `sapply()` and `lapply()`

Instead, we can rewrite it using `sapply()`:
```{r}
sapply(1:3, function(x) x + 1)
sapply(c(1,2,3), function(x) x + 1)
```

* `sapply()` means "simplify apply"
* `lapply()` means "list apply"

Use them to apply a function to each element of a list or vector. The first turns the output into a vector, the latter leaves it as list.

---

### Anonymous functions: the `purrr` way

The `purrr` package (remember the `map()`, `map_dbl()`, etc. functions from last week? they come from this package!) provides another way to write anonymous functions with the **`~.`** syntax. 

Same function as before that adds 1 to the input
```{r}
add <- function(x) {
  x + 1
}
add(c(1,2,3))
```

Rewrite it as anonymous function with purrr (https://purrr.tidyverse.org/reference/map.html)
```{r}
map_dbl(c(1,2,3), ~.x + 1)
```

---

class: inverse, middle

## Practice writing functions (part 2)

Download today's class materials from the website. Complete questions 3, 4, and 5.

---

### Writing more complex functions: example

Consider this simulated data of 4 columns and 10 observations (example taken from Ch. 21 of the 1st Ed of your book):
```{r}
simulated_df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

head(simulated_df, n = 3)
```

<!-- rnorm(10) generates 10 random numbers from a standard normal distribution (mean = 0, sd = 1). -->

---

### Writing more complex functions: example

To compute the mean for each column, we learned we can use a **for loop**:
```{r}
output <- vector("double", length(simulated_df))

for (i in seq_along(simulated_df)) {
  output[i] <- mean(simulated_df[[i]])
}

output
```

---

### Writing more complex functions: example

If you are going to compute the mean for a column frequently, **put your for loop into a function**. This way you can write the code once, and call the function whenever you need it.

--

Put the code from the previous slide into a function and call it `column_mean()`

<!-- copy this code and do it interactively on workbench -->

```{r}
column_mean <- function(df) {
  output <- vector("double", length(df))
  for (i in seq_along(df)) {
    output[i] <- mean(df[[i]])
  }
  return(output)
}

column_mean(simulated_df)
```

---

### Writing more complex functions: example

What if, instead of only computing the mean, we also want to include the median and standard deviation? We **could write a function for each**, replacing  `mean()` with `median()`, `sd()`, etc.

```{r}
column_median <- function(df) {
  output <- vector("double", length(df))
  for (i in seq_along(df)) {
    output[i] <- median(df[[i]])
  }
  return(output)
}
```

```{r}
column_sd <- function(df) {
  output <- vector("double", length(df))
  for (i in seq_along(df)) {
    output[i] <- sd(df[[i]])
  }
  return(output)
}
```

---

### Functions with more than one argument

But this way we copied/pasted the code multiple times, and the differences among these codes are hard to spot!

**How can we rewrite this code more efficiently?** 

**By setting up a function with more than one argument!**

---

### Functions with more than one argument

In this example, the first argument is the dataframe, the second is the operation to perform, which can be another function, like `mean()`, `median()`, `sd()`, etc.

```{r}
column_stats <- function(df, stat) {
  output <- vector("double", length(df))
  for (i in seq_along(df)) {
    output[i] <- stat(df[[i]])
  }
  return(output)
}
```

```{r}
column_stats(simulated_df, median)
column_stats(simulated_df, sd)
```

---

### Functions as a way to organize your code 

Pretty cool right?! **But how many things we can put in a single function?** 
* Ideally, functions should short and do one task. Good habit: break your code into sequential blocks, and write one function for each action (e.g. one function to import data, one to rename variables, one to calculate descriptive statistics, etc.)

* You can chain functions, so that the output of one function is the first argument of the next function: this is called **functions composition**

---

### Functions as a way to organize your code 

Given this vector:
```{r}
v <- c(1:10, NA, NA)
v
```

I want to perform the following two operations, sequentially: remove NAs and calculate the mean; and I set up two functions to perform them.

```{r}
vect_remove_NA <- function(vector){
 return(vector[!is.na(vector)])
}
```

```{r}
vect_mean <- function(vector) {
  return(mean(vector))
}
```

---

### Functions as a way to organize your code 

First, I call the first function, `vect_remove_NA()`, on my vector `v`, and store the output. Then, I pass it to the second function, `vect_mean()`:
```{r}
v_no_missing <- vect_remove_NA(v)
v_mean <- vect_mean(v_no_missing)
v_mean
```

I can do the same thing using pipes:
```{r}
v %>% 
  vect_remove_NA() %>% 
  vect_mean()
```

<!--for more see [here](https://rpubs.com/tjmahr/pipelines_2015) -->

---

### Functions as a way to organize your code

**Takeaway: you can virtually do anything with functions!**

- assign them to variables 
- store them in lists
- pass them as arguments to other functions 
- return them as the result of a function

<!--
https://github.com/annakrystalli/UNAM/blob/master/Functions_in_R.Rmd
-->

---

### Codying style: document your function

Functions should be well-documented! How? 

Each function should include the following:

* a one-sentence description of the function summarizing what it does
* all function's arguments, denoted as `Args:`, with their data types
* a description of the return value, denoted by `Returns:` 

---

### Codying style: document your function

```
column_stats <- function(df, stat) {

  # Computes summary statistics for each column in a data frame
  # Args:
  #   df (tibble): input data
  #   stat (function): a function to compute stats (e.g., mean, median, sd)
  # Returns:
  #   numeric: a vector with the computed statistics for each column of the df
  
  output <- vector("double", length(df))
  for (i in seq_along(df)) {
    output[i] <- stat(df[[i]])
  }
  return(output)
}
```
---

**Example from Google's R Style Guide** [here](https://web.stanford.edu/class/cs109l/unrestricted/resources/google-style.html#:~:text=Function%20Documentation,-Functions%20should%20contain&text=These%20comments%20should%20consist%20of,value%2C%20denoted%20by%20Returns%3A%20.):

```
CalculateSampleCovariance <- function(x, y, verbose = TRUE) {
  # Computes the sample covariance between two vectors.
  # Args:
  #   x: One of two vectors whose sample covariance is to be calculated.
  #   y: The other vector. x and y must have the same length, greater than one,
  #      with no missing values.
  #   verbose: If TRUE, prints sample covariance; if not, not. Default is TRUE.
  # Returns:
  #   The sample covariance between x and y.
  n <- length(x)
  if (n <= 1 || n != length(y)) {
    stop("Arguments x and y have invalid lengths: ",
         length(x), " and ", length(y), ".")
  }
  if (TRUE %in% is.na(x) || TRUE %in% is.na(y)) {
    stop(" Arguments x and y must not have missing values.")
  }
  covariance <- var(x, y)
  if (verbose)
    cat("Covariance = ", round(covariance, 4), ".\n", sep = "")
  return(covariance)
}
```

---

### Writing custom functions in the `tidyverse`

You can write custom functions and use them together with `tidyverse` functions!

For example, you might want to write a custom function to plot a graph in `ggplot` or to perform operations with `dplyr`.

You need the following:
* syntax to write a custom function
* special operators that allows R to evaluate your custom functions correctly: 
   * bang-bang `!!`  with `enquo()` OR
   * curly-curly `{{ }}` (this replaced the former method)

---

### Writing custom functions in the `tidyverse`

Example: create a data frame of age and height of 5 individuals:
```{r}
library(dplyr)

df <- tibble(age = c(10, 12, 13, 14, 10), 
             height = c(110, 140, 155, 170, 130))
df
```

---

### Writing custom functions in the `tidyverse`

Use `summarize()` from `dplyr` to compute mean and standard deviation on the column `age` of this data frame:

```{r}
df %>%
  summarize(mean_age = mean(age),
            sd_age = sd(age))
```

---

### Writing custom functions in the `tidyverse`

Rewrite the previous code using a function and the curly-curly operator: 

```{r}
custom_summary <- function(data, col) {
  data %>%
    summarize(mean_value = mean({{ col }}),
              sd_value = sd({{ col }}))
}

custom_summary(df, age)
```

---

### Writing custom functions in the `tidyverse`

Rewrite it with multiple columns as arguments:

```{r}
custom_summary <- function(data, col1, col2) {
  results <- list(
    summarise(data, 
              mean_value = mean({{ col1 }}),
              sd_value = sd({{ col1 }})),
    summarise(data, 
              mean_value = mean({{ col2 }}),
              sd_value = sd({{ col2 }}))
  )
  # Combine output into a single data frame
  bind_rows(results)
}
custom_summary(df, age, height)
```

---

### Writing custom functions in the `tidyverse`: resources

**Using bang-bang `!!`**

* Chapter ["7 Tidy evaluation"](https://krlmlr.github.io/tidyprog/tidy-evaluation.html): explanation with examples on how and why to use `!!` with `enquo`

* Blog post ["Writing Custom Tidyverse Functions"](https://jonthegeek.com/2018/06/04/writing-custom-tidyverse-functions/): explanation with step-by-step example on how and when to use `!!` with `ensym` and with `enquo`

* Another example [here](http://zevross.com/blog/2018/09/11/writing-efficient-and-streamlined-r-code-with-help-from-the-new-rlang-package/) 

**Using curly-curly `{{ }}`**

* Blog post ["Curly-Curly, the successor of Bang-Bang"](https://www.r-bloggers.com/2019/06/curly-curly-the-successor-of-bang-bang/)

* Another example [here](https://agstats.io/post/writing-r-functions/) (under "Functions and Tidy Evaluation")